---
title: "Differential expression analysis"
subtitle: "THA x SVZ"
author:  |
 | Katia de Paiva Lopes 
 | Raj Lab
 | Department of Neuroscience
 | Icahn School of Medicine at Mount Sinai
 | NYC, New York
date: "`r Sys.Date()`"
output: 
 rmarkdown::html_document:
   theme: united
   highlight: tango
   code_folding: hide
   toc: true
   toc_float: true
   df_print: paged
   smooth_scroll: true
   number_sections: false
   self_contained: true  
---

> Microglia samples.

```{r clean.variables, echo=FALSE}
#This command clean all variables. BE CAREFULL!!! 
rm(list = setdiff(ls(), lsf.str()))
```

```{r load.packages, echo=FALSE, message=FALSE, results='hide'}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
if(!require("limma")) install.packages("limma"); library("limma")
#if(!require("DESeq2")) install.packages("DESeq2"); library("DESeq2")
if(!require("edgeR")) install.packages("edgeR"); library("edgeR")
if(!require("RColorBrewer")) install.packages("RColorBrewer"); library("RColorBrewer")
if(!require("gplots")) install.packages("gplots"); library("gplots")
if(!require("ggplot2")) install.packages("ggplot2"); library("ggplot2")
if(!require("reshape2")) install.packages("reshape2"); library("reshape2")
if(!require("variancePartition")) BiocManager::install("variancePartition"); library("variancePartition")
if(!require("BiocParallel")) install.packages("BiocParallel"); library("BiocParallel")
if(!require("statmod")) install.packages("statmod"); library("statmod")
if(!require("ggpubr")) install.packages("ggpubr"); library("ggpubr")
library('BiocParallel')
library('doParallel')
library(dplyr)
library(tidyr)

# library(devtools)
#install_github("GabrielHoffman/variancePartition")
# BiocManager::version() #Needs to be 3.10

knitr::opts_chunk$set( 
  warning=FALSE,
  message=FALSE,
  results = 'asis',
  error = FALSE,
  tidy = FALSE,
  fig.show = "hold")
```

```{r Helper Functions, echo=FALSE}
createDT <- function(DF, caption="", scrollY=500){
  data <- DT::datatable(DF, caption=caption,
    extensions =  'Buttons',
    options = list( dom = 'Bfrtip', 
                    buttons = c('copy', 'csv', 'excel', 'pdf', 'print'), 
                    scrollY = scrollY, scrollX=T, scrollCollapse = T, paging = F,  
                      columnDefs = list(list(className = 'dt-center', targets = "_all"))
    )
  ) 
   return(data)
}
```

```{r folders, echo=TRUE}
expression_dir = "~/ad-omics_hydra/microglia_omics/expression_tables/added_pilot_314s/expr_4brain_regions/"
work_plots = "~/pd-omics/katia/scripts/GitHub_scripts/glia_omics/3rd_pass_mic_255s/brain_regions/pairwise_dream/"
metadata_path = "~/pd-omics/katia/Microglia/mic_255s/metadata_files/"
```

```{r input}
load(paste0(expression_dir, "Expression_filt_255s.Rdata")) 
# dim(genes_counts_exp_3rd) # 19376   255

load(paste0(metadata_path, "metadata_255filt_eng_29jun2020.Rdata"))
# str(metadata3rd_pass)

metadata <- metadata3rd_pass[, ! colnames(metadata3rd_pass) %in% c("C5", "C6", "C7", "C8", "C9", "C10")] 
# rownames(metadata) = metadata$donor_tissue
```

## Preparing the data for DEG
```{r pick.samples, echo=TRUE}
# metadata$tissue = as.character(metadata$tissue)
samples_reg1 = as.character(metadata$donor_tissue[metadata$tissue %in% "THA"])
samples_reg2 = as.character(metadata$donor_tissue[metadata$tissue %in% "SVZ"])
rownames(metadata) <- metadata$donor_tissue

# Matching the names from the DE analysis to use the same code 
genes_counts4deg = genes_counts_exp_3rd[, colnames(genes_counts_exp_3rd) %in% c(samples_reg1,samples_reg2) ]
metadata4deg = metadata[rownames(metadata) %in% c(samples_reg1,samples_reg2) ,]
# all(rownames(metadata4deg) == colnames(genes_counts4deg)) # # The rownames of the metadata needs to be in the same order as colnames expression table 
# dim(genes_counts4deg) 
# dim(metadata4deg) 
# str(metadata4deg) #shows class for all columns 
metadata4deg$tissue <- as.factor(as.character(metadata4deg$tissue))

table(metadata4deg$tissue)
```

## Dream analysis {.tabset .tabset-fade .tabset-pills} 
```{r dream, echo=TRUE}
params = BiocParallel::MulticoreParam(workers=20, progressbar=T)
register(params)
registerDoParallel(20)

# To include cause of death and C1-C4
metadata4deg$cause_of_death_categories[metadata4deg$cause_of_death_categories %in% NA] <- "Other"
#table(metadata4deg$cause_of_death_categories)

metadata4deg$C1 = metadata4deg$C1 %>% replace_na(median(metadata4deg$C1, na.rm = T))
metadata4deg$C2 = metadata4deg$C2 %>% replace_na(median(metadata4deg$C2, na.rm = T))
metadata4deg$C3 = metadata4deg$C3 %>% replace_na(median(metadata4deg$C3, na.rm = T))
metadata4deg$C4 = metadata4deg$C4 %>% replace_na(median(metadata4deg$C4, na.rm = T))

# Check variance partition version 
# packageVersion("variancePartition")  # Must be 1.17.7

# The variable to be tested should be a fixed effect
form <- ~ sex + (1|donor_id) + age + tissue + (1|cause_of_death_categories) + C1 + C2 + C3 + C4 + picard_pct_mrna_bases + picard_summed_median + picard_pct_ribosomal_bases

# estimate weights using linear mixed model of dream
vobjDream = suppressWarnings( voomWithDreamWeights( genes_counts4deg, form, metadata4deg ) ) # supressing messages because of Biocparallel was generating a lot of messages  
 
# Fit the dream model on each gene
# By default, uses the Satterthwaite approximation for the hypothesis test
fitmm = suppressWarnings (dream( vobjDream, form, metadata4deg )) 

# Examine design matrix
# createDT(fitmm$design, 3)

res_dream <- data.frame(topTable(fitmm, coef='tissueTHA', 
                                 number=nrow(genes_counts4deg), sort.by = "p"), check.names = F)
```

The t-statistics are not directly comparable since they have different degrees of freedom. In order to be able to compare test statistics, we report z.std which is the p-value transformed into a signed z-score. This can be used for downstream analysis.

## DE genes 15% 
```{r dream.deg015, echo=TRUE, results='asis'}
length(which(res_dream$adj.P.Val<0.15))
```

## DE genes 5% 
```{r dream.deg005, echo=TRUE, results='asis'}
length(which(res_dream$adj.P.Val<0.05))
```

## DE genes 0.1% 
```{r dream.deg0001, echo=TRUE, results='asis'}
length(which(res_dream$adj.P.Val<0.01))
```

## Plots from Dream {.tabset .tabset-fade .tabset-pills}
### P Value Distribution {.tabset .tabset-fade .tabset-pills}
```{r design0.pvalue, echo=TRUE, res=300}
res = res_dream
p = ggplot(res, aes(P.Value))
p + geom_density(color="darkblue", fill="lightblue") +
  theme_classic() +
  ggtitle("FDR Distribution")
```

### Fold Change Distribution {.tabset .tabset-fade .tabset-pills}
```{r design0.FC, echo=TRUE, res = 300}
p = ggplot(res, aes(logFC))
p + geom_density(color = "darkblue", fill = "lightblue") + 
  theme_classic() +
  ggtitle("Fold Change Distribution")
```

### MA Plot {.tabset .tabset-fade .tabset-pills}
```{r design0.MA, echo=TRUE, res = 300}
plot.data = res
plot.data$id = rownames(plot.data)
data = data.frame(plot.data)
data$P.Value = -log10(data$P.Value)
data$fifteen = as.factor(abs(data$adj.P.Val < 0.05))

ma = ggplot(data, aes(AveExpr, logFC, color = fifteen))
ma + geom_point() +
  scale_color_manual(values = c("black", "red"), labels = c ("> 0.05", "< 0.05")) +
  labs(title = "MA plot", color = "labels") +
  theme_classic()
  #theme(plot.title = element_text(hjust = 0.5)) + ylim (-10,10) + xlim(-4,22)

```

### Volcano Plot {.tabset .tabset-fade .tabset-pills}
```{r design0.Volcano, echo=TRUE, res = 300}
vp = ggplot(data, aes(logFC, P.Value, color = fifteen))
vp + geom_point() +
  scale_color_manual(values = c("black", "red"), labels = c("> 0.05", "< 0.05")) + 
  labs(title = "Gene Level Volcano Plot", color = "FDR") +
  #theme(plot.title = element_text(hjust = 0.5)) +
  theme_classic() +
  xlim(-10,10) + ylim(0, 20) + ylab("-log10 pvalue")

```

## DEG table for download
```{r design0.lists, echo=TRUE}
## Get conversion table for Gencode 30
gencode_30 = read.table("~/pd-omics/katia/ens.geneid.gencode.v30")
colnames(gencode_30) = c("ensembl","symbol")

res$ensembl = rownames(res)
res_name = merge(res, gencode_30, by="ensembl")
rownames(res_name) = res_name$ensembl
res_name = res_name[order(res_name$adj.P.Val), ]
res_name = res_name[, c("symbol", "logFC", "AveExpr", "t", "P.Value", "adj.P.Val", "z.std")]
   
deg_lists = res_name[which(res_name$adj.P.Val<0.05),]
createDT(res_name)

write.table(res_name, file = paste0(work_plots, "THAxSVZ_dream_3rd.txt"), quote = F, sep = "\t")

```

### Directionality 
LogFC > 0 is up in THA from **Dream**. 
```{r direction5, echo=TRUE, fig.width=10, fig.height=8, res=300}
top_6 = head(deg_lists)
top_6$ensembl = rownames(top_6)
rownames(metadata4deg) = metadata4deg$donor_tissue
genes_voom = genes_counts_voom_3rd[, rownames(metadata4deg)] # Voom 1st pass only for the samples used in this comparison 
gene2check = as.data.frame( genes_voom[rownames(top_6) ,])
gene2check$ensembl = rownames(gene2check)
gene2check = merge(gene2check, top_6[, c("symbol", "ensembl")], by = "ensembl")

gene2check_m = melt(gene2check, id.vars = c("ensembl", "symbol"))
gene2check_charac = merge(gene2check_m, metadata4deg, by.x = "variable", by.y = "donor_tissue")

ggplot(gene2check_charac, aes(x = tissue, y = value, fill = tissue)) +
  geom_boxplot(notch = F, na.rm = T, outlier.color = NA) +
  geom_jitter() +
  theme_classic() + 
  facet_wrap(~symbol, scales = "free_y") +
  ggpubr::stat_compare_means(label = "p.format", label.x.npc = "centre", method = "wilcox.test")
```

```{r session, echo=TRUE}
sessionInfo()
```
