---
title: "Age-related analysis"
subtitle: "Using Dream"
author:  |
 | Katia de Paiva Lopes 
 | Raj Lab
 | Department of Neuroscience
 | Icahn School of Medicine at Mount Sinai
 | NYC, New York
date: "`r Sys.Date()`"
output: 
 rmarkdown::html_document:
   theme: united
   highlight: tango
   code_folding: hide
   toc: true
   toc_float: true
   df_print: paged
   smooth_scroll: true
   number_sections: false
   self_contained: true  
---

> Dream | coef = age.   

```{r clean.variables, echo=FALSE}
#This command clean all variables. BE CAREFULL!!! 
rm(list = setdiff(ls(), lsf.str()))
```

```{r load.packages, echo=FALSE, message=FALSE, results='hide'}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
if(!require("ggeasy")) BiocManager::install("ggeasy"); library("ggeasy")
if(!require("ggsci")) BiocManager::install("ggsci"); library("ggsci")
if(!require("ggplot2")) BiocManager::install("ggplot2"); library("ggplot2")
if(!require("factoextra")) BiocManager::install("factoextra"); library("factoextra")
if(!require("grid")) BiocManager::install("grid"); library("grid")
if(!require("variancePartition")) BiocManager::install("variancePartition"); library("variancePartition")
if(!require("BiocParallel")) install.packages("BiocParallel"); library("BiocParallel")
if(!require("edgeR")) install.packages("edgeR"); library("edgeR")
if(!require("statmod")) install.packages("statmod"); library("statmod")
if(!require("gplots")) install.packages("gplots"); library("gplots")
if(!require("ggpubr")) install.packages("ggpubr"); library("ggpubr")
library(readxl)
library(pheatmap)
library(reshape2)
library(kableExtra)
library(RColorBrewer)
require(gridExtra)
library(ggfortify)
library('doParallel')
library(dplyr)
library(tidyr)

knitr::opts_chunk$set( 
  warning=FALSE,
  message=FALSE,
  results = 'asis',
  error = FALSE,
  tidy = FALSE,
  fig.show = "hold")

#register(MulticoreParam(workers = 10), default = TRUE)
```

```{r Helper Functions, echo=FALSE}
createDT <- function(DF, caption="", scrollY=500){
  data <- DT::datatable(DF, caption=caption,
    extensions =  'Buttons',
    options = list( dom = 'Bfrtip', 
                    buttons = c('copy', 'csv', 'excel', 'pdf', 'print'), 
                    scrollY = scrollY, scrollX=T, scrollCollapse = T, paging = F,  
                      columnDefs = list(list(className = 'dt-center', targets = "_all"))
    )
  ) 
   return(data)
}
```

```{r folders, echo=TRUE}
expression_dir = "~/ad-omics_hydra/microglia_omics/expression_tables/added_pilot_314s/expr_4brain_regions/"
work_plots = "~/pd-omics/katia/scripts/GitHub_scripts/glia_omics/3rd_pass_mic_255s/age_related/age_Dream_3rd/"
work_plots = "~/ad-omics/ricardo/tmp/dream/"
metadata_path = "~/pd-omics/katia/Microglia/mic_255s/metadata_files/"
```

```{r input.data, echo=TRUE}
load(paste0(expression_dir, "Expression_filt_255s.Rdata")) 
# dim(genes_counts_exp_3rd) # 19376   255

load(paste0(metadata_path, "metadata_255filt_eng_29jun2020.Rdata"))
# str(metadata3rd_pass)
```

## Age distribution
### 255 samples 
```{r age_plots, echo=TRUE, fig.width=4, fig.height=3}
metadata <- metadata3rd_pass
ageByDonor = unique(metadata[,c("donor_id", "age", "sex")])

ggplot(ageByDonor, aes(x=age, fill=age)) +
  geom_histogram(bins = 25, colour='black', position = "stack", fill="#5c88da99") +
  labs(x="Age", y="Donors") +
  scale_y_continuous(breaks = (1:20)) +
  scale_x_continuous(breaks=seq(20,120,10)) + 
  theme_classic()

```

## Dream analysis {.tabset .tabset-fade .tabset-pills} 
To use **cause_of_death** in the model, I got the NAs and change it for "Other" category. 

To use C1-C4 I calculated the median for the missing samples. 
```{r dream, echo=TRUE}
params = BiocParallel::MulticoreParam(workers=10, progressbar=T)
register(params)
registerDoParallel(10)

metadata$cause_of_death_categories[metadata$cause_of_death_categories %in% NA] <- "Other"
#table(metadata$cause_of_death_categories)

metadata$C1 = metadata$C1 %>% replace_na(median(metadata$C1, na.rm = T))
metadata$C2 = metadata$C2 %>% replace_na(median(metadata$C2, na.rm = T))
metadata$C3 = metadata$C3 %>% replace_na(median(metadata$C3, na.rm = T))
metadata$C4 = metadata$C4 %>% replace_na(median(metadata$C4, na.rm = T))

# Matching the names from the DE analysis to use the same code 
genes_counts4deg = genes_counts_exp_3rd
metadata4deg = metadata
# all(colnames(genes_counts_exp_3rd) == metadata$donor_tissue) # Check the order of columns - TRUE 

# The dream model operates directly on the results of voom. 
# The only change compared to the standard limma workflow is to replace lmFit with dream. 

# Check variance partition version 
# packageVersion("variancePartition")  # Must be 1.17.7


# The variable to be tested should be a fixed effect
form <- ~ sex + (1|donor_id) + age + tissue + (1|cause_of_death_categories) + C1 + C2 + C3 + C4 + picard_pct_mrna_bases + picard_summed_median + picard_pct_ribosomal_bases

# estimate weights using linear mixed model of dream
vobjDream = suppressWarnings( voomWithDreamWeights( genes_counts4deg, form, metadata4deg ) ) # supressing messages because of Biocparallel was generating a lot of messages  
 
# Fit the dream model on each gene
# By default, uses the Satterthwaite approximation for the hypothesis test
fitmm = suppressWarnings (dream( vobjDream, form, metadata4deg )) 

# Examine design matrix
# createDT(fitmm$design, 3)

res_age <- data.frame(topTable(fitmm, coef='age', 
                                 number=nrow(genes_counts4deg), sort.by = "p"), check.names = F)

```

## Genes related with age
### adj.P.Val<0.001
```{r design0.deg1, echo=TRUE, results='asis'}
length(which(res_age$adj.P.Val<0.001))
```

### adj.P.Val<0.01
```{r design0.deg2, echo=TRUE, results='asis'}
length(which(res_age$adj.P.Val<0.01))
```

### adj.P.Val<0.05
```{r design0.deg3, echo=TRUE, results='asis'}
length(which(res_age$adj.P.Val<0.05))
```

```{r design0.lists, echo=TRUE}
res = res_age
## Get conversion table for Gencode 30
gencode_30 = read.table("~/pd-omics/katia/ens.geneid.gencode.v30")
colnames(gencode_30) = c("ensembl","symbol")

res$ensembl = rownames(res)
res_name = merge(res, gencode_30, by="ensembl")
rownames(res_name) = res_name$ensembl
res_name = res_name[order(res_name$adj.P.Val), ]
res_name = res_name[, c("symbol", "logFC", "AveExpr", "t", "P.Value", "adj.P.Val")]
   
#deg_lists = res_name[which(res_name$adj.P.Val<0.001),]

write.table(res_name, file = paste0(work_plots, "List_age_dream_255s.txt"), quote = F, sep = "\t")
# write.table(deg_lists, file = paste0(work_plots, "List_age_0.001_dream.txt"), quote = F, sep = "\t")

createDT(res_name)
```

## PCA voom 
The voom here comes from the Dream object. 
```{r pca_voom, echo=TRUE, fig.width=6, fig.height=4}
genes_voom = as.data.frame(vobjDream$E) 
res.pca = prcomp(t(genes_voom)) 

autoplot(res.pca, data = metadata4deg, colour = 'age') +
  scale_colour_viridis_c() +
  theme_classic()

```

I'm applying the removeBatchEffect function to get the residuals. 
```{r remove_batch, echo=TRUE}
allResiduals <- removeBatchEffect(x = genes_voom, 
                                  batch = metadata4deg$tissue, 
                                  batch2 = paste(metadata4deg$sex, metadata4deg$cause_of_death_categories),
                                 # design = model.matrix(~ age, data = metadata4deg), #force to not regress age. It didn't change the PCA. 
                                  covariates = as.matrix(metadata4deg[, c("picard_pct_mrna_bases", "picard_summed_median", "picard_pct_ribosomal_bases", "C1","C2","C3","C4" )]))

```

## PCA after removeBatchEffect
```{r pca_noBatch, echo=TRUE, fig.width=6, fig.height=4}
res.pca = prcomp(t(allResiduals)) 

autoplot(res.pca, data = metadata4deg, colour = 'age') +
  scale_colour_viridis_c() +
  theme_classic()
```

## All samples (n = 255)
Ordered by age! 

I've applied the function scale to transform the data into z-score. 
```{r sel_age, echo=TRUE, fig.width=12, fig.height=20}
age_genes = res_name[res$adj.P.Val<0.001, ]
age_genes = age_genes[order(age_genes$adj.P.Val) ,]
top_age_genes = age_genes[1:100 ,]

residuals_matrix = as.data.frame(allResiduals)
residuals_matrix$ensembl = rownames(residuals_matrix)
residuals_matrix = merge(residuals_matrix, gencode_30, by="ensembl")

#Remove duplicates 
residuals_matrix <- residuals_matrix[! duplicated(residuals_matrix$symbol),]
rownames(residuals_matrix) = residuals_matrix$symbol
residuals_matrix$ensembl = NULL
residuals_matrix$symbol = NULL
# dim(residuals_matrix) #19342   255

ourexpr_age_res = residuals_matrix[rownames(residuals_matrix) %in% top_age_genes$symbol , ] # RESIDUALS TABLE 
col_age = as.data.frame(metadata4deg[c("age")] ,)
rownames(col_age) = metadata4deg$donor_tissue
summary(col_age)
ourexpr_age_zscore = t(scale(t(ourexpr_age_res))) # The scale is by column. I want the scale by row so I need to transpose the matrix. Then, I transposed again for the plot! 
ourexpr_age_zscore = as.data.frame(ourexpr_age_zscore)

all(colnames(ourexpr_age_zscore) == metadata4deg$Donor_tissue) # Check the order of columns - TRUE 
my.breaks <- seq(-3, 3, length.out = 100)

ourexpr_age_zscore_sorted = ourexpr_age_zscore[, order(col_age)]

# Function to cluster the columns beautifully! 
# callback = function(hc, mat){
#     sv = svd(t(mat))$v[,1]
#     dend = reorder(as.dendrogram(hc), wts = sv)
#     as.hclust(dend)
# }

pheatmap::pheatmap(ourexpr_age_zscore_sorted, 
                   annotation_col = col_age, 
                   show_colnames = F,
                   breaks = my.breaks,
                   cluster_cols = F,
                   #clustering_callback = callback,
                   clustering_method = "ward.D2") 

```

```{r}
sessionInfo()
```



