---
title: "Age-related analysis"
subtitle: "Using Dream"
author:  |
 | Katia de Paiva Lopes 
 | Ricardo A. Vialle
 | Raj Lab
 | Department of Neuroscience
 | Icahn School of Medicine at Mount Sinai
 | NYC, New York
date: "`r Sys.Date()`"
output: 
 rmarkdown::html_document:
   theme: united
   highlight: tango
   code_folding: hide
   toc: true
   toc_float: true
   df_print: paged
   smooth_scroll: true
   number_sections: false
   self_contained: true  
---

> Dream | coef = age.   

```{r clean.variables, echo=FALSE}
#This command clean all variables. BE CAREFULL!!! 
rm(list = setdiff(ls(), lsf.str()))
```

```{r load.packages, echo=FALSE, message=FALSE, results='hide'}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
if(!require("ggeasy")) BiocManager::install("ggeasy"); library("ggeasy")
if(!require("ggsci")) BiocManager::install("ggsci"); library("ggsci")
if(!require("ggplot2")) BiocManager::install("ggplot2"); library("ggplot2")
if(!require("factoextra")) BiocManager::install("factoextra"); library("factoextra")
if(!require("grid")) BiocManager::install("grid"); library("grid")
if(!require("variancePartition")) BiocManager::install("variancePartition"); library("variancePartition")
if(!require("BiocParallel")) install.packages("BiocParallel"); library("BiocParallel")
if(!require("edgeR")) BiocManager::install("edgeR"); library("edgeR")
if(!require("statmod")) install.packages("statmod"); library("statmod")
if(!require("gplots")) install.packages("gplots"); library("gplots")
if(!require("ggpubr")) install.packages("ggpubr"); library("ggpubr")
library(readxl)
library(pheatmap)
library(reshape2)
library(kableExtra)
library(RColorBrewer)
require(gridExtra)
library(ggfortify)
library('doParallel')
library(dplyr)
library(tidyr)
library(tidyverse)
library(circlize)
library(pacman)

knitr::opts_chunk$set( 
  warning=FALSE,
  message=FALSE,
  results = 'asis',
  error = FALSE,
  tidy = FALSE,
  fig.show = "hold")

#register(MulticoreParam(workers = 10), default = TRUE)
```

```{r Helper Functions, echo=FALSE}
createDT <- function(DF, caption="", scrollY=500){
  data <- DT::datatable(DF, caption=caption,
    extensions =  'Buttons',
    options = list( dom = 'Bfrtip', 
                    buttons = c('copy', 'csv', 'excel', 'pdf', 'print'), 
                    scrollY = scrollY, scrollX=T, scrollCollapse = T, paging = F,  
                      columnDefs = list(list(className = 'dt-center', targets = "_all"))
    )
  ) 
   return(data)
}
```

```{r folders, echo=TRUE}
expression_dir = "~/ad-omics_hydra/microglia_omics/expression_tables/added_pilot_314s/expr_4brain_regions/"
#work_plots = "~/pd-omics/katia/scripts/GitHub_scripts/glia_omics/3rd_pass_mic_255s/age_related/age_Dream_3rd/"
work_plots = "~/ad-omics/ricardo/MiGA/age_interaction/age_interaction/"
metadata_path = "~/pd-omics/katia/Microglia/mic_255s/metadata_files/"
```

```{r input.data, echo=TRUE}
load(paste0(expression_dir, "Expression_filt_255s.Rdata"))
# dim(genes_counts_exp_3rd) # 19376   255

load(paste0(metadata_path, "metadata_255filt_eng_29jun2020.Rdata"))
# str(metadata3rd_pass)
```

## Age distribution
### 255 samples 
```{r age_plots, echo=TRUE, fig.width=3, fig.height=2.5}
metadata <- metadata3rd_pass
ageByDonor = unique(metadata[,c("donor_id", "age", "sex")])

ggplot(ageByDonor, aes(x=age, fill=age)) +
  geom_histogram(bins = 25, colour='black', position = "stack", fill="#5c88da99") +
  labs(x="Age", y="Donors") +
  scale_y_continuous(breaks = seq(0,20,2.5)) +
  scale_x_continuous(breaks=seq(20,120,10)) + 
  theme_classic()
```

## Dream analysis {.tabset .tabset-fade .tabset-pills} 
To use **cause_of_death** in the model, I got the NAs and change it for "Other" category. 

To use C1-C4 I calculated the median for the missing samples. 
```{r dream, echo=TRUE, eval=F}
params = BiocParallel::MulticoreParam(workers=10, progressbar=T)
register(params)
registerDoParallel(10)

metadata$cause_of_death_categories[metadata$cause_of_death_categories %in% NA] <- "Other"
#table(metadata$cause_of_death_categories)

metadata$C1 = metadata$C1 %>% replace_na(median(metadata$C1, na.rm = T))
metadata$C2 = metadata$C2 %>% replace_na(median(metadata$C2, na.rm = T))
metadata$C3 = metadata$C3 %>% replace_na(median(metadata$C3, na.rm = T))
metadata$C4 = metadata$C4 %>% replace_na(median(metadata$C4, na.rm = T))

# Matching the names from the DE analysis to use the same code 
genes_counts4deg = genes_counts_exp_3rd
metadata4deg = metadata
# all(colnames(genes_counts_exp_3rd) == metadata$donor_tissue) # Check the order of columns - TRUE 

# The dream model operates directly on the results of voom. 
# The only change compared to the standard limma workflow is to replace lmFit with dream. 

# Check variance partition version 
# packageVersion("variancePartition")  # Must be 1.17.7

# The variable to be tested should be a fixed effect
form <- ~ sex + (1|donor_id) + (1|cause_of_death_categories) + C1 + C2 + C3 + C4 + picard_pct_mrna_bases + picard_summed_median + picard_pct_ribosomal_bases + age * tissue

# estimate weights using linear mixed model of dream
vobjDream = suppressWarnings( voomWithDreamWeights( genes_counts4deg, form, metadata4deg ) ) # supressing messages because of Biocparallel was generating a lot of messages  
 
# Fit the dream model on each gene
# By default, uses the Satterthwaite approximation for the hypothesis test
metadata4deg$tissue = relevel(metadata4deg$tissue, ref = "MFG")
fitmm_MFGref = suppressWarnings (dream( vobjDream, form, metadata4deg )) 
metadata4deg$tissue = relevel(metadata4deg$tissue, ref = "STG")
fitmm_STGref = suppressWarnings (dream( vobjDream, form, metadata4deg )) 
metadata4deg$tissue = relevel(metadata4deg$tissue, ref = "SVZ")
fitmm_SVZref = suppressWarnings (dream( vobjDream, form, metadata4deg ))
metadata4deg$tissue = relevel(metadata4deg$tissue, ref = "THA")
fitmm_THAref = suppressWarnings (dream( vobjDream, form, metadata4deg )) 

save(vobjDream, fitmm_MFGref,fitmm_STGref,fitmm_SVZref,fitmm_THAref, metadata4deg, genes_counts4deg,  file = paste0(work_plots,"/age_interaction_dreamObj.RData"))
```

## DE genes with interaction term

### adj.P.Val<0.01
```{r echo=TRUE, dpi=300, fig.width=3, fig.height=2.2}
load(paste0(work_plots,"/age_interaction_dreamObj.RData"))

MFG_0.01 = colSums(as.data.frame(unclass(summary(decideTests(fitmm_MFGref, adjust.method = "BH", p.value = 0.01))))[c(1,3),])
MFG_0.05 = colSums(as.data.frame(unclass(summary(decideTests(fitmm_MFGref, adjust.method = "BH", p.value = 0.05))))[c(1,3),])
MFG_0.10 = colSums(as.data.frame(unclass(summary(decideTests(fitmm_MFGref, adjust.method = "BH", p.value = 0.10))))[c(1,3),])

STG_0.01 = colSums(as.data.frame(unclass(summary(decideTests(fitmm_STGref, adjust.method = "BH", p.value = 0.01))))[c(1,3),])
STG_0.05 = colSums(as.data.frame(unclass(summary(decideTests(fitmm_STGref, adjust.method = "BH", p.value = 0.05))))[c(1,3),])
STG_0.10 = colSums(as.data.frame(unclass(summary(decideTests(fitmm_STGref, adjust.method = "BH", p.value = 0.10))))[c(1,3),])

SVZ_0.01 = colSums(as.data.frame(unclass(summary(decideTests(fitmm_SVZref, adjust.method = "BH", p.value = 0.01))))[c(1,3),])
SVZ_0.05 = colSums(as.data.frame(unclass(summary(decideTests(fitmm_SVZref, adjust.method = "BH", p.value = 0.05))))[c(1,3),])
SVZ_0.10 = colSums(as.data.frame(unclass(summary(decideTests(fitmm_SVZref, adjust.method = "BH", p.value = 0.10))))[c(1,3),])

THA_0.01 = colSums(as.data.frame(unclass(summary(decideTests(fitmm_THAref, adjust.method = "BH", p.value = 0.01))))[c(1,3),])
THA_0.05 = colSums(as.data.frame(unclass(summary(decideTests(fitmm_THAref, adjust.method = "BH", p.value = 0.05))))[c(1,3),])
THA_0.10 = colSums(as.data.frame(unclass(summary(decideTests(fitmm_THAref, adjust.method = "BH", p.value = 0.10))))[c(1,3),])


df_res_0.01 = data.frame(MFG = c(NA,MFG_0.01["age:tissueSTG"],MFG_0.01["age:tissueSVZ"],MFG_0.01["age:tissueTHA"]),
                         STG = c(STG_0.01["age:tissueMFG"],NA,STG_0.01["age:tissueSVZ"],STG_0.01["age:tissueTHA"]),
                         SVZ = c(SVZ_0.01["age:tissueMFG"],SVZ_0.01["age:tissueSTG"],NA,SVZ_0.01["age:tissueTHA"]),
                         THA = c(THA_0.01["age:tissueMFG"],THA_0.01["age:tissueSTG"],THA_0.01["age:tissueSVZ"],NA))
rownames(df_res_0.01) = c("MFG","STG","SVZ","THA")
p_load(ComplexHeatmap)
ComplexHeatmap::Heatmap(as.matrix(df_res_0.01), 
                        name = "DE FDR<1%",
                        cell_fun = function(j, i, x, y, width, height, fill) {
                            grid.text(sprintf("%d", df_res_0.01[i, j]), x, y, gp = gpar(fontsize = 10))}, 
                        cluster_columns = F, 
                        cluster_rows = F, 
                        col = colorRamp2(c(min(df_res_0.01,na.rm = T),max(df_res_0.01,na.rm = T)), c("white", "#CB454A")), rect_gp = gpar(col = "white", lwd = 2) )
```

### adj.P.Val<0.05
```{r echo=TRUE,  dpi=300, fig.width=3, fig.height=2.2}
df_res_0.05 = data.frame(MFG = c(NA,MFG_0.05["age:tissueSTG"],MFG_0.05["age:tissueSVZ"],MFG_0.05["age:tissueTHA"]),
                         STG = c(STG_0.05["age:tissueMFG"],NA,STG_0.05["age:tissueSVZ"],STG_0.05["age:tissueTHA"]),
                         SVZ = c(SVZ_0.05["age:tissueMFG"],SVZ_0.05["age:tissueSTG"],NA,SVZ_0.05["age:tissueTHA"]),
                         THA = c(THA_0.05["age:tissueMFG"],THA_0.05["age:tissueSTG"],THA_0.05["age:tissueSVZ"],NA))
rownames(df_res_0.05) = c("MFG","STG","SVZ","THA")

ComplexHeatmap::Heatmap(as.matrix(df_res_0.05), 
                        name = "DE FDR<5%",
                        cell_fun = function(j, i, x, y, width, height, fill) {
                            grid.text(sprintf("%d", df_res_0.05[i, j]), x, y, gp = gpar(fontsize = 10))}, 
                        cluster_columns = F, 
                        cluster_rows = F, 
                        col = colorRamp2(c(min(df_res_0.05,na.rm = T),max(df_res_0.05,na.rm = T)), c("white", "#CB454A")), rect_gp = gpar(col = "white", lwd = 2) )
```

### adj.P.Val<0.10
```{r echo=TRUE, dpi=300, fig.width=3, fig.height=2.2}
df_res_0.10 = data.frame(MFG = c(NA,MFG_0.10["age:tissueSTG"],MFG_0.10["age:tissueSVZ"],MFG_0.10["age:tissueTHA"]),
                         STG = c(STG_0.10["age:tissueMFG"],NA,STG_0.10["age:tissueSVZ"],STG_0.10["age:tissueTHA"]),
                         SVZ = c(SVZ_0.10["age:tissueMFG"],SVZ_0.10["age:tissueSTG"],NA,SVZ_0.10["age:tissueTHA"]),
                         THA = c(THA_0.10["age:tissueMFG"],THA_0.10["age:tissueSTG"],THA_0.10["age:tissueSVZ"],NA))
rownames(df_res_0.10) = c("MFG","STG","SVZ","THA")

ComplexHeatmap::Heatmap(as.matrix(df_res_0.10), 
                        name = "DE FDR<10%",
                        cell_fun = function(j, i, x, y, width, height, fill) {
                            grid.text(sprintf("%d", df_res_0.10[i, j]), x, y, gp = gpar(fontsize = 10))}, 
                        cluster_columns = F, 
                        cluster_rows = F, 
                        col = colorRamp2(c(min(df_res_0.10,na.rm = T),max(df_res_0.10,na.rm = T)), c("white", "#CB454A")), rect_gp = gpar(col = "white", lwd = 2) )
```

## Show some DE genes in the interaction term

### List of age related genes of the interaction between SVZ and MFG

```{r echo=TRUE}
res_age <- data.frame(topTable(fitmm_MFGref, coef='age:tissueSVZ', number=nrow(genes_counts4deg)), check.names = F)

## Get conversion table for Gencode 30
gencode_30 = read.table("~/pd-omics/katia/ens.geneid.gencode.v30", header = T)
colnames(gencode_30) = c("ensembl","symbol")

res_age$ensembl = rownames(res_age)
res_name = merge(res_age, gencode_30, by="ensembl")
rownames(res_name) = res_name$ensembl
res_name = res_name[order(res_name$adj.P.Val), ]
res_name = res_name[, c("symbol", "logFC", "AveExpr", "t", "P.Value", "adj.P.Val")]
   
createDT(res_name)
```

### List of age related genes of the interaction between SVZ and THA

```{r echo=TRUE}
res_age <- data.frame(topTable(fitmm_THAref, coef='age:tissueSVZ', number=nrow(genes_counts4deg)), check.names = F)

## Get conversion table for Gencode 30
gencode_30 = read.table("~/pd-omics/katia/ens.geneid.gencode.v30", header = T)
colnames(gencode_30) = c("ensembl","symbol")

res_age$ensembl = rownames(res_age)
res_name = merge(res_age, gencode_30, by="ensembl")
rownames(res_name) = res_name$ensembl
res_name = res_name[order(res_name$adj.P.Val), ]
res_name = res_name[, c("symbol", "logFC", "AveExpr", "t", "P.Value", "adj.P.Val")]
   
createDT(res_name)
```

### List of all interaction age-tissue genes at FDR<5% for all 6 possible comparisons

```{r echo=TRUE}
res_age_MFG_SVZ <- data.frame(topTable(fitmm_MFGref, coef='age:tissueSVZ', number=nrow(genes_counts4deg)), check.names = F) %>% rownames_to_column("ensembl") %>% mutate(comparison = "MFG_SVZ") 
res_age_MFG_THA <- data.frame(topTable(fitmm_MFGref, coef='age:tissueTHA', number=nrow(genes_counts4deg)), check.names = F) %>% rownames_to_column("ensembl") %>% mutate(comparison = "MFG_THA")
res_age_MFG_STG <- data.frame(topTable(fitmm_MFGref, coef='age:tissueSTG', number=nrow(genes_counts4deg)), check.names = F) %>% rownames_to_column("ensembl") %>% mutate(comparison = "MFG_STG")
res_age_STG_SVZ <- data.frame(topTable(fitmm_STGref, coef='age:tissueSVZ', number=nrow(genes_counts4deg)), check.names = F) %>% rownames_to_column("ensembl") %>% mutate(comparison = "STG_SVZ")
res_age_STG_THA <- data.frame(topTable(fitmm_STGref, coef='age:tissueTHA', number=nrow(genes_counts4deg)), check.names = F) %>% rownames_to_column("ensembl") %>% mutate(comparison = "STG_THA")
res_age_SVZ_THA <- data.frame(topTable(fitmm_SVZref, coef='age:tissueTHA', number=nrow(genes_counts4deg)), check.names = F) %>% rownames_to_column("ensembl") %>% mutate(comparison = "SVZ_THA")

res_age = rbind(res_age_MFG_SVZ,res_age_MFG_THA,res_age_MFG_STG,res_age_STG_SVZ,res_age_STG_THA,res_age_SVZ_THA)

## Get conversion table for Gencode 30
gencode_30 = read.table("~/pd-omics/katia/ens.geneid.gencode.v30", header = T)
colnames(gencode_30) = c("ensembl","symbol")

res_name = merge(res_age, gencode_30, by="ensembl")
res_name = res_name[order(res_name$adj.P.Val), ]

#write.table(x = res_name, file = paste0(work_plots,"all_comparisons.txt"), quote = F, sep = "\t", row.names = F, col.names = T)
createDT(res_name %>% filter(adj.P.Val<0.05))
```
### List with significant interaction at FDR 5% in any comparison (group by gene)

91 genes demonstrated age-region relationships after fitting an interaction term model (FDR < 0.05)

```{r warning=FALSE, message=FALSE, results='hide', fig.keep='all'}
res_signif = res_name %>% filter(adj.P.Val<0.05) %>% group_by(ensembl,symbol) %>% summarise(comparisons = paste(sort(comparison), collapse = ";"), n = n())

res_signif %>% arrange(-n)
```

### Plot top-5 genes (MFG vs SVZ)

Expression in log10(TPM)

```{r warning=FALSE, message=FALSE, results='hide', fig.keep='all', dpi=300, fig.width=10, fig.height=10}
metadata4deg$tissue = factor(metadata4deg$tissue, levels = c("MFG","STG","SVZ","THA"))
# After running the limma you can extract the results using the topTable function.
# Here I pick the gene name from the top DE gene
top5Gene = rownames(topTable(fitmm_MFGref, coef = "age:tissueSVZ"))[1:5]

# Now get the expression of the gene after running regressing covariates
geneExp = t(log10(genes_tpm_exp_3rd[top5Gene,]+1))
rownames(gencode_30) = gencode_30$ensembl
colnames(geneExp) = gencode_30[colnames(geneExp),"symbol"]
geneExp = as.data.frame(geneExp)

# Add the expression as another column in the metadata table (samples in both tables should be in the same order!)
meta_tmp = cbind(metadata4deg, geneExp)
meta_tmp_m = melt(meta_tmp, measure.vars = c(colnames(geneExp)))

lm_eqn <- function(df, y, x){
  formula = as.formula(sprintf('%s ~ %s', y, x))
  m <- lm(formula, data=df);
  # formating the values into a summary string to print out
  # ~ give some space, but equal size and comma need to be quoted
  eq <- substitute(italic(target) == a + b %.% italic(input)*","~~italic(r)^2~"="~r2*","~~p~"="~italic(pvalue), 
                   list(target = y,
                        input = x,
                        a = format(as.vector(coef(m)[1]), digits = 2), 
                        b = format(as.vector(coef(m)[2]), digits = 2), 
                        r2 = format(summary(m)$r.squared, digits = 3),
                        # getting the pvalue is painful
                        pvalue = format(summary(m)$coefficients[2,'Pr(>|t|)'], digits=1)
                   )
  )
  as.character(as.expression(eq));                 
}

# Plot the expression of the gene.
library(ggpmisc)
ggplot(meta_tmp_m, aes(x = age, y = value, color = tissue)) + 
  geom_point() +
  scale_color_futurama() + 
  easy_labs(x = "Age", y = expression(paste(log[10],"(TPM+1)"))) +
  geom_smooth(method='lm', alpha = .3, se = F) +
  stat_poly_eq(aes(label = paste(..rr.label..)), 
               rr.digits = 2, 
               label.x = 0.1, label.y = .98,
               formula = y ~ x, parse = TRUE, size = 4) +
  facet_grid(variable ~ tissue, scales = "free") + 
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
  easy_remove_legend()  
#ggsave(paste0(work_plots,"/top5_agerelated_TPM.pdf"), width = 10, height = 10)
```

Same thing but with residuals (z-scored)

```{r warning=FALSE, message=FALSE, results='hide', fig.keep='all', dpi=300, fig.width=10, fig.height=10}
plot_residuals_per_age <- function(genes){
  # Get the residuals after adjusting for covariates
  residuals = removeBatchEffect(vobjDream$E, # data after running voom
      batch = metadata4deg$tissue, # add one categorical covariate
      batch2 = paste(metadata4deg$sex,"_",metadata4deg$cause_of_death_categories), # add a second categorical covariate ( if you want )
      # here you can force the regression to NOT remove effects of some variable of interest
      #design = model.matrix(~ + age + tissue, data = metadata4deg),
      # add the column with continuous covariates
      covariates = metadata4deg %>% dplyr::select(picard_pct_mrna_bases, picard_summed_median, picard_pct_ribosomal_bases, C1 , C2 , C3 , C4))
  
  residuals_zscore = t(scale(t(residuals)))
  
  # Now get the expression of the gene after running regressing covariates
  geneExp = t(residuals_zscore[genes,]+1)
  rownames(gencode_30) = gencode_30$ensembl
  colnames(geneExp) = gencode_30[colnames(geneExp),"symbol"]
  geneExp = as.data.frame(geneExp)
  
  # Add the expression as another column in the metadata table (samples in both tables should be in the same order!)
  meta_tmp = cbind(metadata4deg, geneExp)
  meta_tmp_m = melt(meta_tmp, measure.vars = c(colnames(geneExp)))
  
  # Plot the expression of the gene.
  library(ggpmisc)
  ggplot(meta_tmp_m, aes(x = age, y = value, color = tissue)) + 
    geom_point() +
    scale_color_futurama() + 
    easy_labs(x = "Age", y = expression(paste("Residual expression (Z-score)"))) +
    geom_smooth(method='lm', alpha = .3, se = F) +
    stat_poly_eq(aes(label = paste(..adj.rr.label..)), 
                 rr.digits = 2, 
                 label.x = 0.1, label.y = .98,
                 formula = y ~ x, parse = TRUE, size = 4) +
    facet_grid(variable ~ tissue, scales = "free") + 
    theme_bw() + 
    theme(panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank()) +
    easy_remove_legend()
}
top5Gene = rownames(topTable(fitmm_MFGref, coef = "age:tissueSVZ"))[1:5]
plot_residuals_per_age(top5Gene)
```

## Get gene lists from interaction results (paper review)

A set of them had significant increased (n = 16) or decreased (n = 11) expression during aging only in MFG and STG compared with SVZ.

Selecting only genes with significant interaction between MFG vs SVZ and STG vs SVZ.

```{r warning=FALSE, message=FALSE}
mfg_stg_vs_svz = res_signif[res_signif$comparisons=="MFG_SVZ;STG_SVZ",]
mfg_stg_vs_svz_up <- res_name[res_name$ensembl %in% mfg_stg_vs_svz$ensembl,] %>% filter(adj.P.Val<0.05, logFC>0) %>% group_by(ensembl,symbol) %>% summarise(comparisons = paste(sort(comparison), collapse = ";"), n = n())
mfg_stg_vs_svz_up$direction = "Up in SVZ"
mfg_stg_vs_svz_down <- res_name[res_name$ensembl %in% mfg_stg_vs_svz$ensembl,] %>% filter(adj.P.Val<0.05, logFC<0) %>% group_by(ensembl,symbol) %>% summarise(comparisons = paste(sort(comparison), collapse = ";"), n = n())
mfg_stg_vs_svz_down$direction = "Down in SVZ"
createDT(rbind(mfg_stg_vs_svz_down, mfg_stg_vs_svz_up))
```
 
Checking the plots (16 genes up in MFG and STG but down in SVZ)
 
```{r warning=FALSE, message=FALSE, results='hide', fig.keep='all', dpi=300, fig.width=10, fig.height=16*2}
plot_residuals_per_age(mfg_stg_vs_svz_down$ensembl)
```

Checking the plots (11 genes down in MFG and STG but up in SVZ)

```{r warning=FALSE, message=FALSE, results='hide', fig.keep='all', dpi=300, fig.width=10, fig.height=11*2}
plot_residuals_per_age(mfg_stg_vs_svz_up$ensembl)
```

25 genes (e.g. MRC1, CD24) changed specifically in SVZ and not in other regions.

```{r warning=FALSE, message=FALSE}
svz_vs_all = res_signif[grep("SVZ", res_signif$comparisons),]

# Get the residuals after adjusting for covariates
residuals = removeBatchEffect(vobjDream$E, # data after running voom
  batch = metadata4deg$tissue, # add one categorical covariate
  batch2 = paste(metadata4deg$sex,"_",metadata4deg$cause_of_death_categories), # add a second categorical covariate ( if you want )
  # here you can force the regression to NOT remove effects of some variable of interest
  #design = model.matrix(~ + age + tissue, data = metadata4deg),
  # add the column with continuous covariates
  covariates = metadata4deg %>% dplyr::select(picard_pct_mrna_bases, picard_summed_median, picard_pct_ribosomal_bases, C1 , C2 , C3 , C4))

residuals_zscore = t(scale(t(residuals)))
residuals_zscore = residuals

# Now get the expression of the gene after running regressing covariates
geneExp = t(residuals_zscore[unique(svz_vs_all$ensembl),]+1)
rownames(gencode_30) = gencode_30$ensembl
colnames(geneExp) = gencode_30[colnames(geneExp),"symbol"]
geneExp = as.data.frame(geneExp)

# Add the expression as another column in the metadata table (samples in both tables should be in the same order!)
meta_tmp = cbind(metadata4deg, geneExp)
meta_tmp_m = melt(meta_tmp, measure.vars = c(colnames(geneExp)))

# Run a linear regression between age and residual expression  
age_regression <- meta_tmp_m %>% group_by(variable,tissue) %>% 
  do( as.data.frame(coef(summary(glm(value ~ age, data=.)))) [2,])
rownames(age_regression) = paste0(age_regression$variable, "_", age_regression$tissue)

age_regression2 <- meta_tmp_m %>% group_by(variable,tissue) %>% 
  do( data.frame(R2 = summary(lm(value ~ age, data=.))$adj.r.squared) )
rownames(age_regression2) = paste0(age_regression2$variable, "_", age_regression2$tissue)

age_regression3 = bind_cols(age_regression, age_regression2[,"R2",drop=F])

betas = age_regression3 %>% select(variable, tissue, Estimate) %>% pivot_wider(values_from = Estimate, names_from = tissue) %>% column_to_rownames(var = "variable")
pvalues = age_regression3 %>% select(variable, tissue, `Pr(>|t|)`) %>% pivot_wider(values_from = `Pr(>|t|)`, names_from = tissue) %>% column_to_rownames(var = "variable")
rsquares = age_regression3 %>% select(variable, tissue, R2) %>% pivot_wider(values_from = R2, names_from = tissue) %>% column_to_rownames(var = "variable")

#with_effect_in_SVZ = abs(rsquares$SVZ) > rowMeans(abs(rsquares[,c("MFG","STG","THA")]))
p_load(matrixStats)

#with_effect_in_SVZ = abs(rsquares$SVZ) > 5*rowMeans(as.matrix(abs(rsquares[,c("MFG","STG","THA")])))
#rsquares[with_effect_in_SVZ,]

with_effect_in_SVZ = abs(rsquares$SVZ) > 3*rowIQRs(as.matrix(abs(rsquares[,c("MFG","STG","THA")])))

createDT(rsquares[with_effect_in_SVZ,])
```

Checking the plots
 
```{r warning=FALSE, message=FALSE, results='hide', fig.keep='all', dpi=300, fig.width=10, fig.height=35*2}
plot_residuals_per_age(gencode_30[gencode_30$symbol %in% rownames(rsquares[with_effect_in_SVZ,]), "ensembl"])
```

# SessionInfo

```{r}
sessionInfo()
```
