---
title: "Plotting the Effect Sizes"
subtitle: " "
author:  |
 | Katia de Paiva Lopes 
 | Raj Lab
 | Department of Neuroscience
 | Icahn School of Medicine at Mount Sinai
 | NYC, New York
date: "`r Sys.Date()`"
output: 
 rmarkdown::html_document:
   theme: united
   highlight: tango
   code_folding: hide
   toc: true
   toc_float: true
   df_print: paged
   smooth_scroll: true
   number_sections: false
   self_contained: true  
---

```{r clean.variables, echo=FALSE}
#This command clean all variables. BE CAREFULL!!! 
rm(list = setdiff(ls(), lsf.str()))
```

```{r load.packages, echo=FALSE, message=FALSE, results='hide'}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
if(!require("RColorBrewer")) install.packages("RColorBrewer"); library("RColorBrewer")
if(!require("gplots")) install.packages("gplots"); library("gplots")
if(!require("ggplot2")) install.packages("ggplot2"); library("ggplot2")
if(!require("reshape2")) install.packages("reshape2"); library("reshape2")
library(ggpubr)
library(ggrepel)
library(venn)
library(ggsci)
library("readxl")
library(ggeasy)
library(dplyr)
library(ggpmisc)
library(tidyverse)

knitr::opts_chunk$set( 
  warning=FALSE,
  message=FALSE,
  results = 'asis',
  error = FALSE,
  tidy = FALSE,
  fig.show = "hold")
```

```{r Helper Functions, echo=FALSE}
createDT <- function(DF, caption="", scrollY=500){
  data <- DT::datatable(DF, caption=caption,
    extensions =  'Buttons',
    options = list( dom = 'Bfrtip', 
                    buttons = c('copy', 'csv', 'excel', 'pdf', 'print'), 
                    scrollY = scrollY, scrollX=T, scrollCollapse = T, paging = F,  
                      columnDefs = list(list(className = 'dt-center', targets = "_all"))
    )
  ) 
   return(data)
}
```

```{r folders, echo=TRUE}
work_plots = "/Users/katia/OneDrive/Documentos/MountSinai/Projects/Microglia/Figures4paper/Fig_age/age_effect_size/"
lists2compare = "~/pd-omics/katia/scripts/GitHub_scripts/glia_omics/3rd_pass_mic_255s/age_related/age_by_region/"
inter_terms = "~/pd-omics/katia/scripts/GitHub_scripts/glia_omics/3rd_pass_mic_255s/age_interaction/"
```

```{r pd_de, echo=TRUE}
deg_list_1 = read.table(file = paste0(lists2compare, "List_age_limma_MFG.txt"), header = T)
deg_list_1 = deg_list_1[order(deg_list_1$adj.P.Val) ,]
#length(which(deg_list_1$adj.P.Val<0.05)) 
deg_list_1$study = "age_MFG"
deg_list_1$ensembl = rownames(deg_list_1)

deg_list_2 = read.table(file = paste0(lists2compare, "List_age_limma_STG.txt"), header = T)
deg_list_2 = deg_list_2[order(deg_list_2$adj.P.Val) ,]
# length(which(deg_list_2$adj.P.Val<0.05)) 
deg_list_2$study = "age_STG"
deg_list_2$ensembl = rownames(deg_list_2)

deg_list_3 = read.table(file = paste0(lists2compare, "List_age_limma_THA.txt"), header = T)
deg_list_3 = deg_list_3[order(deg_list_3$adj.P.Val) ,]
# length(which(deg_list_3$adj.P.Val<0.05)) 
deg_list_3$study = "age_THA"
deg_list_3$ensembl = rownames(deg_list_3)

deg_list_4 = read.table(file = paste0(lists2compare, "List_age_limma_SVZ.txt"), header = T)
deg_list_4 = deg_list_4[order(deg_list_4$adj.P.Val) ,]
# length(which(deg_list_4$adj.P.Val<0.05)) 
deg_list_4$study = "age_SVZ"
deg_list_4$ensembl = rownames(deg_list_4)

inter_terms_list = read.table(file = paste0(inter_terms, "interaction_list_77g.txt"), header = T)

```

## Age - all genes 
Beta values of all genes. 
```{r effect_size1, fig.width=6, fig.height=4, dpi=300}

deg_list_all = deg_list_1[,c("ensembl","symbol","logFC","study")] %>% 
  left_join(deg_list_2[,c("ensembl","logFC")], by = "ensembl") %>%
  left_join(deg_list_3[,c("ensembl","logFC")], by = "ensembl") %>%
  left_join(deg_list_4[,c("ensembl","logFC")], by = "ensembl")
colnames(deg_list_all) = c("ensembl","symbol","logFC_MFG","MFG","logFC_STG","logFC_THA","logFC_SVZ")

genes_m = melt(deg_list_all, id.vars = c("ensembl","symbol","logFC_MFG","MFG"))

genes_m$variable = gsub("(.*)_(.*)", "\\2", genes_m$variable)
genes_m$variable = factor(genes_m$variable, levels = c("STG","THA","SVZ"))

#pdf(file = paste0(work_plots, "beta_age_all.pdf"), width = 6, height = 4)
ggplot2::ggplot(genes_m, aes(x = logFC_MFG, y = value, color = variable)) + 
  geom_point(alpha = .5) +
  scale_color_manual(values = c("#C71000FF", "#8A4198FF", "#008EA0FF")) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "black") + 
  stat_smooth(method = "lm", se=F) + # Add Regression Line 
  stat_regline_equation(aes(label = ..adj.rr.label..), show.legend = F)  + # Add R-Square
#  stat_regline_equation(aes(label = ..rr.label..))  +
  easy_labs(x = expression(paste("MFG age-related (", beta,")")), y = expression(paste("age-related (", beta, ")"))) +
  easy_add_legend_title("Region") +
  theme_classic()
#dev.off()
```

## Age - filtered 1
Shows only the genes with FDR < 0.05 in MFG.

X = beta in MFG. 
Y = beta in the other brain regions (by color). 
```{r effect_size2, fig.width=6, fig.height=4, dpi=300}
genes_m_filt = genes_m[genes_m$ensembl %in% deg_list_1$ensembl[deg_list_1$adj.P.Val < 0.05], ]

#pdf(file = paste0(work_plots, "beta_age_005_MFG.pdf"), width = 6, height = 4)
ggplot2::ggplot(genes_m_filt, aes(x = logFC_MFG, y = value, color = variable)) + 
  geom_point() +
  scale_color_manual(values = c("#C71000FF", "#8A4198FF", "#008EA0FF")) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "black") + 
  stat_smooth(method = "lm", se=F) + # Add Regression Line 
  stat_regline_equation(aes(label = ..adj.rr.label..), show.legend = F)  + # Add R-Square
#  stat_regline_equation(aes(label = ..rr.label..))  +
  easy_labs(x = expression(paste("MFG age-related (", beta,")")), y = expression(paste("age-related (", beta, ")"))) +
  easy_add_legend_title("Region") +
  theme_classic()
#dev.off()
```

## Age - filtered 2
Shows only the genes with FDR < 0.05 in at least one region + the interaction terms genes (77 genes from MFGxSVZ).

X = beta in MFG. 
Y = beta in the other brain regions (by color). 
```{r effect_size3, fig.width=6, fig.height=4, dpi=300}
genes_m_filt = genes_m[genes_m$ensembl %in% unique(c(deg_list_1$ensembl[deg_list_1$adj.P.Val < 0.05],
                                                     deg_list_2$ensembl[deg_list_2$adj.P.Val < 0.05],
                                                     deg_list_3$ensembl[deg_list_3$adj.P.Val < 0.05],
                                                     deg_list_4$ensembl[deg_list_4$adj.P.Val < 0.05],
                                                   as.character(inter_terms_list$ensembl))), ]

symbol4label = c("MS4A6A", "MERTK", "ANXA1", "MRC1", "LYVE1") 
genes_m_filt4lab = genes_m_filt
genes_m_filt4lab$overlap = genes_m_filt4lab$symbol
genes_m_filt4lab$overlap[(!genes_m_filt4lab$overlap %in% symbol4label) | genes_m_filt4lab$variable!="SVZ"] = ""

# pdf(file = paste0(work_plots, "beta_age_sign_005_oneregion_label_interac.pdf"), width = 6, height =4)
ggplot2::ggplot(genes_m_filt4lab, aes(x = logFC_MFG, y = value, color = variable)) + 
  geom_point() +
  scale_color_manual(values = c("#C71000FF", "#8A4198FF", "#008EA0FF")) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "lightgrey") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "lightgrey") + 
  stat_smooth(method = "lm", se=F) + # Add Regression Line +
  stat_poly_eq(formula = y ~ x, aes(label = paste(..adj.rr.label..,..p.value.label..,sep = "*`,`~")), parse=TRUE) +
 # stat_regline_equation(aes(label = paste(..adj.rr.label..)), show.legend = F)  + # Add R-Square
#  stat_regline_equation(aes(label = ..rr.label..))  +
  #geom_text_repel(aes(label = overlap),color="black",force = 4) + # Descoment to show the symbol
  geom_label_repel(aes(label = overlap), size = 3, color="black", box.padding = 0.4, label.size = NA, fill = alpha(c("white"),0.5)) + # Descoment to show the symbol
  easy_labs(x = expression(paste("MFG age-related (", beta,")")), y = expression(paste("age-related (", beta, ")"))) +
  easy_add_legend_title("Region") +
  theme_classic()
#dev.off()
```

## Age - filtered 3
Shows the top 100 DE genes in each region (lists might overlap).

X = beta in MFG. 
Y = beta in the other brain regions (by color). 
```{r effect_size5, fig.width=6, fig.height=4, dpi=300}
genes_m_filt = genes_m[genes_m$ensembl %in% unique(c(deg_list_1$ensembl[1:100],
                                                     deg_list_2$ensembl[1:100],
                                                     deg_list_3$ensembl[1:100],
                                                     deg_list_4$ensembl[1:100])), ]

#pdf(file = paste0(work_plots, "beta_age_sign_top100_oneregion.pdf"), width = 6, height = 4)
ggplot2::ggplot(genes_m_filt, aes(x = logFC_MFG, y = value, color = variable)) + 
  geom_point() +
  scale_color_manual(values = c("#C71000FF", "#8A4198FF", "#008EA0FF")) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "black") + 
  geom_text_repel(data = genes_m_filt[genes_m_filt$logFC_MFG*genes_m_filt$value < 0, ], aes(label = symbol)) + # Descoment to show the symbol
  stat_smooth(method = "lm", se=F) + # Add Regression Line 
  stat_regline_equation(aes(label = ..adj.rr.label..), show.legend = F)  + # Add R-Square
#  stat_regline_equation(aes(label = ..rr.label..))  +
  easy_labs(x = expression(paste("MFG age-related (", beta,")")), y = expression(paste("age-related (", beta, ")"))) +
  easy_add_legend_title("Region") +
  theme_classic()
#dev.off()
```

```{r}
sessionInfo()
```

