---
title: "Looking at covariates"
subtitle: ""
author:  |
 | Katia de Paiva Lopes 
 | Raj Lab
 | Department of Neuroscience
 | Icahn School of Medicine at Mount Sinai
 | NYC, New York
date: "`r Sys.Date()`"
output: 
 rmarkdown::html_document:
   theme: united
   highlight: tango
   code_folding: show
   toc: true
   toc_float: true
   df_print: paged
   smooth_scroll: true
   number_sections: false
   self_contained: true  
---

> 255 samples (from 314) | 
> Different brain regions from the same donor | 
> Multi disease cohort. 

Previous R script: 01_expression_metadata_4reg.R by Katia Lopes. 

```{r clean.variables, echo=FALSE}
#This command clean all variables. BE CAREFULL!!! 
rm(list = setdiff(ls(), lsf.str()))
```

```{r load.packages, echo=FALSE, message=FALSE, results='hide'}
if(!require("variancePartition")) install.packages("variancePartition"); library("variancePartition")
if(!require("limma")) install.packages("limma"); library("limma")
if(!require("stats")) install.packages("stats"); library("stats")
if(!require("ggfortify")) install.packages("ggfortify"); library("ggfortify")
if(!require("gplots")) install.packages("gplots"); library("gplots")
if(!require("factoextra")) install.packages("factoextra"); library("factoextra")
if(!require("pamr")) BiocManager::install("pamr"); library("pamr")
if(!require("DESeq2")) BiocManager::install("DESeq2"); library("DESeq2")
if(!require("edgeR")) BiocManager::install("edgeR"); library("edgeR")
if(!require("broom")) BiocManager::install("broom"); library("broom")
if(!require("tidyverse")) BiocManager::install("tidyverse"); library("tidyverse")


knitr::opts_chunk$set( 
  warning=FALSE,
  message=FALSE,
  results = 'asis',
  error = FALSE,
  tidy = FALSE,
  fig.show = "hold")
```

```{r Helper Functions, echo=FALSE}
createDT <- function(DF, caption="", scrollY=500){
  data <- DT::datatable(DF, caption=caption,
    extensions =  'Buttons',
    options = list( dom = 'Bfrtip', 
                    buttons = c('copy', 'csv', 'excel', 'pdf', 'print'), 
                    scrollY = scrollY, scrollX=T, scrollCollapse = T, paging = F,  
                      columnDefs = list(list(className = 'dt-center', targets = "_all"))
    )
  ) 
   return(data)
}
```

```{r folders, echo=TRUE}
work_dir = "~/ad-omics/ricardo/tmp/3rd_pass_mic_258s/"
expression_dir = "~/ad-omics_hydra/microglia_omics/expression_tables/added_pilot_314s/expr_4brain_regions/"
metadata_path = "~/pd-omics/katia/Microglia/mic_255s/metadata_files/"
```

## Input: metadata
This metadata contains biological and technical covariates. 
```{r input.metadata, echo=TRUE}
load(paste0(metadata_path, "metadata_255filt_eng_29jun2020.Rdata"))
# str(metadata3rd_pass)

metadata <- metadata3rd_pass[, ! colnames(metadata3rd_pass) %in% c("C5", "C6", "C7", "C8", "C9", "C10")] 
rownames(metadata) = metadata$donor_tissue
```

## HM: correlating the covariates 
<div style="text-align: justify"> 
"Evaluating the correlation between variables in a important part in interpreting variancePartition results. When comparing two **continuous** variables, Pearson
correlation is widely used. But variancePartition includes **categorical** variables in the model as well. In order to accommodate the correlation between a continuous and a categorical variable, or two categorical variables we used canonical correlation analysis. **Canonical Correlation Analysis (CCA)** is similar to correlation between two vectors, except that CCA can accommodate matricies as well." Gabriel Hoffman. 
</div>
```{r hm_variance_partition, echo=TRUE, fig.width=10, fig.height=8, dpi=300}

# These variables have zero variance so cannot be analyzed:
#  infection_2weeks, glucocorticosteroids, batch

metadata_alt1 = metadata %>% dplyr::select(-glucocorticosteroids) # These variables have zero variance so cannot be analyzed: glucocorticosteroids
metadata_alt2 = metadata_alt1 %>% dplyr::select(-infection_2weeks) # These variables have zero variance so cannot be analyzed: infection_2weeks
metadata_alt = metadata_alt2 %>% dplyr::select(-batch) # These variables have zero variance so cannot be analyzed: batch

# names(metadata_alt) = tolower(names(metadata_alt))

# Transforms characters columns in factors columns 
indx <- sapply(metadata_alt, is.character)
metadata_alt[indx] <- lapply(metadata_alt[indx], function(x) as.factor(x))
metadata_alt$ph = as.numeric(as.character(metadata_alt$ph))
#str(metadata_alt) #shows class for all columns 
form <- as.formula(paste("", paste(colnames(metadata_alt), collapse=" + "), sep=" ~ ")) #form contains the design 
C = canCorPairs(form, metadata_alt)

# pdf(paste0(work_dir, "figures_pdf/canonical_hm.pdf"), width = 10, height = 8)
plotCorrMatrix(C, margins = c(12,18))
# dev.off()

```

```{r tissues.donor, eval=FALSE}
### Number of tissues by donor 
tissuesBydonor = rowSums(table(metadata_alt[,c("donor_id","tissue")]))
createDT(as.data.frame(tissuesBydonor))
```

### Donors with 01 tissue 
```{r donors_1tissue, echo=TRUE}
sum(rowSums(table(metadata_alt[,c("donor_id","tissue")]))==1)
```

```{r donor.id_1tissue, echo=TRUE}
donor1sample = names(which(rowSums(table(metadata_alt[,c("donor_id","tissue")]))>1))
donor1sample
```
## Input expression 
```{r expression, echo=TRUE}
load(paste0(expression_dir, "Expression_filt_255s.Rdata")) 
# dim(genes_counts_exp_3rd) # 19376   255
```

## VP: 255 samples
The covariates for the formula was chosen based on canonical correlation. We need to be careful to interpret this plot because here, we are using all samples together: same individual with different brain regions. 

If the covariate have NAs in the column, we can't fit a model for Limma or Dream. The following covariates have NAs: **lane, benzodiapezines, opiates, autoimmune_diseases, smoking, infection_2weeks, alcohol_dependence_daily_use, ph, C1-C4 from genotyping, suicide_attempts and cause_of_death_categories**. 
```{r vp_alls, echo=TRUE, fig.width=8, fig.height=6, res=300 }
library(doParallel)
cl <- makeCluster(10)
registerDoParallel(cl)

# For categorical covariates you need to include the number 1. Example: (1|Status) + (1|Plate)
metadata_alt$fastqc_total_sequences = scale(metadata_alt$fastqc_total_sequences)
metadata_alt$star_uniquely_mapped = scale(metadata_alt$star_uniquely_mapped)
metadata_alt$featurecounts_assigned = scale(metadata_alt$featurecounts_assigned)

# form <- ~ (1|donor_id) + age + (1|tissue) + (1|lane) + (1|main_diagnosis) + (1|sex) + pmd_minutes + picard_pct_mrna_bases + picard_summed_mean + picard_pct_pf_reads_aligned + ph + picard_pct_ribosomal_bases + (1|cause_of_death_categories) + C1 + C2 + C3 + C4 + (1|suicide_attempts) + fastqc_total_sequences + (1|opiates) + (1|autoimmune_diseases) + star_uniquely_mapped + fastqc_percent_fails + featurecounts_assigned 

form <- ~ (1|donor_id) + age + (1|tissue) + (1|lane) + (1|main_diagnosis) + (1|sex) + pmd_minutes + picard_pct_mrna_bases + picard_summed_mean + picard_pct_pf_reads_aligned + ph + picard_pct_ribosomal_bases + (1|cause_of_death_categories) + C1 + C2 + C3 + C4 

varPart_tx <- suppressWarnings( fitExtractVarPartModel( genes_counts_voom_3rd, form, metadata_alt) )
vp <- sortCols( varPart_tx )
# plotPercentBars( vp[1:10,] ) Not applicable for dataset with different tissues from the same individual 

save(varPart_tx, file = paste0(work_dir, "/vp_files/varPart_tx_255s.RData"))
plotVarPart(vp) +
  ggeasy::easy_rotate_x_labels(angle = 70, side = c("right"))
```

## VP: By tissue (255 samples) {.tabset .tabset-fade .tabset-pills} 

We have tried to run including +0 in the formula (Vignette, page 19) but it didn't works because we have some donors with only one tissue. So, we can't fit a model for that. However, we can try to run separately by tissue. 

```{r split.tissue, echo=TRUE}
metadata_GFM = metadata_alt[metadata_alt$tissue=="MFG",]
voom_GFM = genes_counts_voom_3rd[ , as.character(metadata_GFM$donor_tissue)]

metadata_GTS = metadata_alt[metadata_alt$tissue=="STG",]
voom_GTS = genes_counts_voom_3rd[ , as.character(metadata_GTS$donor_tissue)]

metadata_SVZ = metadata_alt[metadata_alt$tissue=="SVZ",] 
voom_SVZ = genes_counts_voom_3rd[ , as.character(metadata_SVZ$donor_tissue)]

metadata_THA = metadata_alt[metadata_alt$tissue=="THA",] 
voom_THA = genes_counts_voom_3rd[ , as.character(metadata_THA$donor_tissue)]

save(metadata_alt, metadata_GFM, metadata_GTS, metadata_SVZ, metadata_THA,
             voom_GFM, voom_GTS, voom_SVZ, voom_THA, file = paste0(work_dir, "metadata_exprBytissue_255s.Rdata"))
```

```{r vp_formBytissue, echo=TRUE }

form <- ~ age + (1|lane) + (1|main_diagnosis) + (1|sex) + pmd_minutes + picard_pct_mrna_bases + picard_summed_mean + picard_pct_pf_reads_aligned + ph + picard_pct_ribosomal_bases + (1|cause_of_death_categories) + C1 + C2 + C3 + C4 

```

### MFG - Medial Frontal Gyrus {.tabset .tabset-fade .tabset-pills}
```{r vp_gfm, echo=TRUE, fig.width=8, fig.height=6, res=300 }
varPart_tx <- suppressWarnings( fitExtractVarPartModel( voom_GFM, form, metadata_GFM) )
vp <- sortCols( varPart_tx )
save(varPart_tx, file = paste0(work_dir, "vp_files/varPart_tx_MFG.RData"))

plotVarPart(vp)+
  ggeasy::easy_rotate_x_labels(angle = 70, side = c("right"))
```

### STG - Superior Temporal Gyrus {.tabset .tabset-fade .tabset-pills}
```{r vp_gts, echo=TRUE, fig.width=8, fig.height=6, res=300 }
varPart_tx <- suppressWarnings( fitExtractVarPartModel( voom_GTS, form, metadata_GTS) )
vp <- sortCols( varPart_tx )
save(varPart_tx, file = paste0(work_dir, "vp_files/varPart_tx_STG.RData"))

plotVarPart(vp)+
  ggeasy::easy_rotate_x_labels(angle = 70, side = c("right"))
```

### SVZ - Subventricular zone {.tabset .tabset-fade .tabset-pills}
```{r vp_svz, echo=TRUE, fig.width=8, fig.height=6, res=300 }
varPart_tx <- suppressWarnings( fitExtractVarPartModel( voom_SVZ, form, metadata_SVZ) )
vp <- sortCols( varPart_tx )
save(varPart_tx, file = paste0(work_dir, "vp_files/varPart_tx_SVZ.RData"))

plotVarPart(vp) +
  ggeasy::easy_rotate_x_labels(angle = 70, side = c("right"))
```

### THA - Thalamus {.tabset .tabset-fade .tabset-pills}
```{r vp_tha, echo=TRUE, fig.width=8, fig.height=6, res=300 }
varPart_tx <- suppressWarnings( fitExtractVarPartModel( voom_THA, form, metadata_THA) )
vp <- sortCols( varPart_tx )
save(varPart_tx, file = paste0(work_dir, "vp_files/varPart_tx_THA.RData"))

plotVarPart(vp)+
  ggeasy::easy_rotate_x_labels(angle = 70, side = c("right"))
```

## VP: 216 samples, European only 
```{r vp_216s, echo=TRUE, fig.width=8, fig.height=6, dpi=300}
eur_samples = read.table("~/pd-omics/katia/Microglia/mic_255s/samples_eqtl_90d.txt", header = F, stringsAsFactors = T)
colnames(eur_samples) = c("eur_samples")

metadata_alt_eur = metadata_alt[metadata_alt$donor_id %in% eur_samples$eur_samples ,]
genes_counts_voom_3rd_eur = genes_counts_voom_3rd[, colnames(genes_counts_voom_3rd) %in% metadata_alt_eur$donor_tissue]

form <- ~ (1|donor_id) + age + (1|tissue) + (1|lane) + (1|main_diagnosis) + (1|sex) + pmd_minutes + picard_pct_mrna_bases + picard_summed_mean + picard_pct_pf_reads_aligned + ph + picard_pct_ribosomal_bases + (1|cause_of_death_categories) + C1 + C2 + C3 + C4 

varPart_tx <- suppressWarnings( fitExtractVarPartModel( genes_counts_voom_3rd_eur, form, metadata_alt_eur) )
vp <- sortCols( varPart_tx )
# plotPercentBars( vp[1:10,] ) Not applicable for dataset with different tissues from the same individual 

save(varPart_tx, file = paste0(work_dir, "/vp_files/varPart_tx_eur_216s.RData"))
plotVarPart(vp) +
  ggeasy::easy_rotate_x_labels(angle = 70, side = c("right"))

```

```{r session, echo=TRUE}
sessionInfo()
```

