---
title: "Exploratory plots"
subtitle: "PCAs"
author:  |
 | Katia de Paiva Lopes 
 | Raj Lab
 | Department of Neuroscience
 | Icahn School of Medicine at Mount Sinai
 | NYC, New York
date: "`r Sys.Date()`"
output: 
 rmarkdown::html_document:
   theme: united
   highlight: tango
   code_folding: hide
   toc: true
   toc_float: true
   df_print: paged
   smooth_scroll: true
   number_sections: false
   self_contained: true  
---

> 255 samples | 
> Different brain regions from the same donor | 
> Multi disease cohort. 

```{r clean.variables, echo=FALSE}
#This command clean all variables. BE CAREFULL!!! 
rm(list = setdiff(ls(), lsf.str()))
```

```{r load.packages, echo=FALSE, message=FALSE, results='hide'}

library(knitr)
library(dplyr)
library(kableExtra)
library(ggplot2)
library(RColorBrewer)
require(grid)
require(gridExtra)
library(ggfortify)
library(ggsci)
library(factoextra)
library(dplyr)
library(tidyr)
library(ggeasy)
library(ggpubr)
library("edgeR")

knitr::opts_chunk$set( 
  warning=FALSE,
  message=FALSE,
  results = 'asis',
  error = FALSE,
  tidy = FALSE,
  fig.show = "hold")
```

```{r Helper Functions, echo=FALSE}
createDT <- function(DF, caption="", scrollY=500){
  data <- DT::datatable(DF, caption=caption,
    extensions =  'Buttons',
    options = list( dom = 'Bfrtip', 
                    buttons = c('copy', 'csv', 'excel', 'pdf', 'print'), 
                    scrollY = scrollY, scrollX=T, scrollCollapse = T, paging = F,  
                      columnDefs = list(list(className = 'dt-center', targets = "_all"))
    )
  ) 
   return(data)
}
```


```{r folders, echo=TRUE}
expression_dir = "~/ad-omics_hydra/microglia_omics/expression_tables/added_pilot_314s/expr_4brain_regions/"
work_plots = "~/pd-omics/katia/scripts/GitHub_scripts/glia_omics/3rd_pass_mic_255s/"
metadata_path = "~/pd-omics/katia/Microglia/mic_255s/metadata_files/"
plots4paper = "/Users/katia/OneDrive/Documentos/MountSinai/Projects/Microglia/Figures4paper/"
sup_fig = "/Users/katia/OneDrive/Documentos/MountSinai/Projects/Microglia/Suplementary_material/expl_plots/"
```

```{r input.data, echo=TRUE}
load(paste0(expression_dir, "Expression_filt_255s.Rdata")) 
# dim(genes_counts_exp_3rd) # 19376   255

load(paste0(metadata_path, "metadata_255filt_eng_29jun2020.Rdata"))
# str(metadata3rd_pass)
metadata <- metadata3rd_pass
# To use **cause_of_death** in the model, I got the NAs and change it for "Other" category. 
#To use C1-C4 I calculated the median for the missing samples. 

metadata$cause_of_death_categories[metadata$cause_of_death_categories %in% NA] <- "Other"
#table(metadata$cause_of_death_categories)

metadata$C1 = metadata$C1 %>% replace_na(median(metadata$C1, na.rm = T))
metadata$C2 = metadata$C2 %>% replace_na(median(metadata$C2, na.rm = T))
metadata$C3 = metadata$C3 %>% replace_na(median(metadata$C3, na.rm = T))
metadata$C4 = metadata$C4 %>% replace_na(median(metadata$C4, na.rm = T))

```

## PCAs
PCs 1 and 2. 
Voom without correction. 
```{r pca.diverse, echo=TRUE, fig.width=12, fig.height=8, dpi=300}
res.pca = prcomp(t(genes_counts_voom_3rd)) 

g1 <- autoplot(res.pca, data = metadata, colour = 'tissue') +
  scale_color_futurama() +
  easy_add_legend_title("Region") +
theme_classic()

g2 <- autoplot(res.pca, data = metadata, colour = 'cause_of_death_categories') +
  scale_colour_igv() +
  easy_add_legend_title("Cause of death") +
  theme_classic() 

g3 <- autoplot(res.pca, data = metadata, colour = 'main_diagnosis') +
  scale_colour_manual(values = c("darkviolet", "orange", "green", "darkblue", "lightblue", "red", "pink", "yellow", "#5C88DAFF")) +
  easy_add_legend_title("Diagnosis") +
  theme_classic()

g4 <- autoplot(res.pca, data = metadata, colour = 'pmd_minutes') +
  scale_color_viridis_c() +
  easy_add_legend_title("PMD minutes") +
  theme_classic()

g5 <- autoplot(res.pca, data = metadata, colour = 'batch') +
theme_classic() 

# pdf(paste0(plots4paper, "SUPP_PCAs_voom_255s.pdf"), width=12, height=8)
#grid.arrange(g1,g2,g3,g4, ncol = 2)
ggarrange(g1,g2,g3,g4, ncol=2, nrow = 2, labels = c("a", "b", "c", "d"))
# dev.off()
# plot(g2)
```

## Remove BatchEffects 
### Region (n = 255) 

a = Voom without correction. b = After removeBatchEffect without regressing out tissue. 

```{r pca2, fig.width=10, fig.height=4, dpi=300}
allResiduals <- removeBatchEffect(x = genes_counts_voom_3rd, 
                                  batch = metadata$sex, 
                                  batch2 = metadata$cause_of_death_categories,
                                  design = model.matrix(~ tissue, data = metadata), #force to not regress tissue. 
                                  covariates = as.matrix(metadata[, c("picard_pct_mrna_bases", "picard_summed_median", "picard_pct_ribosomal_bases", "C1","C2","C3","C4" )]))

res.pca = prcomp(t(allResiduals)) 

# pdf(paste0(plots4paper, "PCA_255s_residuals.pdf"), width=4, height=3)
g6 <- autoplot(res.pca, data = metadata, colour = 'tissue') +
  scale_colour_futurama() +
  easy_add_legend_title("Region") +
 # scale_colour_viridis_c() +
  theme_classic()
# dev.off()

# pdf(paste0(plots4paper, "PCAs_255s_voomxresiduals.pdf"), width=10, height=4)
ggarrange(g1,g6, labels = c("a", "b"))
# dev.off()
```

## Remove BatchEffects 
### Age (n = 255) 

a = Voom without correction. b = After removeBatchEffect without regressing out age -> colored by age. 

```{r pca_age, fig.width=10, fig.height=4, dpi=300}

# MFG = #FF6F00FF
# STG = #C71000FF
# SVZ = #008EA0FF
# THA = #8A4198FF

res.pca = prcomp(t(genes_counts_voom_3rd)) 

k1 <- autoplot(res.pca, data = metadata, colour = 'age') +
 # scale_color_gradient2(low = "#008EA0FF", midpoint = 60, mid = "#FF6F00FF", high = "#8A4198FF", na.value = NA ) +
  scale_color_viridis_c(option = "magma") + 
  easy_add_legend_title("Age") +
  theme_classic()

allResiduals <- removeBatchEffect(x = genes_counts_voom_3rd, 
                                  batch = metadata$sex, 
                                  batch2 = metadata$cause_of_death_categories,
                                  design = model.matrix(~ tissue, data = metadata), #force to not regress tissue. 
                                  covariates = as.matrix(metadata[, c("picard_pct_mrna_bases", "picard_summed_median", "picard_pct_ribosomal_bases", "C1","C2","C3","C4" )]))

res.pca = prcomp(t(allResiduals)) 

k2 <- autoplot(res.pca, data = metadata, colour = 'age') +
  scale_color_viridis_c(option = "magma") + 
  easy_add_legend_title("Age") +
  theme_classic()

# pdf(paste0(plots4paper, "PCAs_255s_voomxresiduals_Age.pdf"), width=10, height=4)
ggarrange(k1,k2, labels = c("a", "b"))
# dev.off()

# pdf(paste0(plots4paper, "PCAs_255s_Age.pdf"), width=3, height=2)
plot(k2)
# dev.off()

```


### (n = 112)
Only the donors who share the four main brain regions.  
a = Voom without correction. b = After removeBatchEffect without regressing out tissue. 

Covariates removed: sex, cause_of_death_categories, picard_pct_mrna_bases, picard_summed_median, picard_pct_ribosomal_bases and C1-C4. Without regressing out tissue.
```{r pca_112s, fig.width=10, fig.height=4, dpi=300}
metadata_numb <- metadata
donors_4tissues <- metadata_numb %>% group_by(donor_id) %>% summarise(n=n()) %>% filter(n==4)
metadata_numb <- metadata[metadata$donor_id %in% donors_4tissues$donor_id ,]
metadata_numb$cause_of_death_categories <- as.factor(as.character(metadata_numb$cause_of_death_categories))

genes_voom_numb <- genes_counts_voom_3rd[, colnames(genes_counts_voom_3rd) %in% metadata_numb$donor_tissue]
# all(colnames(genes_voom_numb) == metadata_numb$donor_tissue) # Check the order 
# dim(genes_voom_numb)

res.pca = prcomp(t(genes_voom_numb)) 

g7 <- autoplot(res.pca, data = metadata_numb, colour = 'tissue') +
  scale_colour_futurama() +
  easy_add_legend_title("Region") +
  theme_classic()

allResiduals <- removeBatchEffect(x = genes_voom_numb, 
                                  batch = metadata_numb$sex, 
                                  batch2 = metadata_numb$cause_of_death_categories,
                                 # design = model.matrix(~ tissue, data = metadata_numb), #force to not regress tissue. 
                                  covariates = as.matrix(metadata_numb[, c("picard_pct_mrna_bases", "picard_summed_median", "picard_pct_ribosomal_bases", "C1","C2","C3","C4" )]))

res.pca = prcomp(t(allResiduals)) 

g8 <- autoplot(res.pca, data = metadata_numb, colour = 'tissue') +
  scale_colour_futurama() +
  easy_add_legend_title("Region") +
  theme_classic()

# pdf(paste0(plots4paper, "PCA_112s_voomxresiduals.pdf"), width=10, height=4)
ggarrange(g7,g8, labels = c("a", "b"))
# dev.off()

# pdf(paste0(plots4paper, "PCA_112s_Region_panel.pdf"), width=3, height=2)
plot(g8)
# dev.off()
```

### Donors are numbers
PCA with donors who share the 4 main brain tissues. 112 samples from 28 donors. 

a = Voom without correction. b = After removeBatchEffect without regressing out tissue. 
```{r numb, fig.width=10, fig.height=4, dpi=300}
metadata_numb$donor_numb = as.numeric(as.factor(metadata_numb$donor_id)) # Gera números por donor! 

metadata_numb$tissue_sm = as.character(metadata_numb$tissue)
metadata_numb$tissue_sm[metadata_numb$tissue_sm == "MFG"] <- "m"
metadata_numb$tissue_sm[metadata_numb$tissue_sm == "STG"] <- "s"
metadata_numb$tissue_sm[metadata_numb$tissue_sm == "THA"] <- "t"
metadata_numb$tissue_sm[metadata_numb$tissue_sm == "SVZ"] <- "z"
metadata_numb$tissue_sm <- as.factor(metadata_numb$tissue_sm)

metadata_numb$donor_numb2 <- paste0(metadata_numb$donor_numb, metadata_numb$tissue_sm)
rownames(metadata_numb) <- metadata_numb$donor_numb2

genes_voom_numb <- genes_counts_voom_3rd[, colnames(genes_counts_voom_3rd) %in% metadata_numb$donor_tissue]
# all(colnames(genes_voom_numb) == metadata_numb$donor_tissue) # Check the order 
colnames(genes_voom_numb) <- rownames(metadata_numb)

res.pca = prcomp(t(genes_voom_numb)) 

g9 <- autoplot(res.pca, data = metadata_numb, colour = 'tissue', label=T, shape = F) +
  scale_colour_futurama() +
  easy_add_legend_title("Region") +
 # scale_colour_viridis_c() +
  theme_classic()
 
############# Using residuals 
residuals_numb <- allResiduals[, colnames(allResiduals) %in% metadata_numb$donor_tissue ] #Calculated in the later chunck for 112 samples
# all(colnames(residuals_numb) == metadata_numb$donor_tissue)
colnames(residuals_numb) <- rownames(metadata_numb)

res.pca = prcomp(t(allResiduals)) 

g10 <- autoplot(res.pca, data = metadata_numb, colour = 'tissue', label=T, shape = F) +
  scale_colour_futurama() +
  easy_add_legend_title("Region") +
 # scale_colour_viridis_c() +
  theme_classic()

#pdf(paste0(plots4paper, "PCA_112s_voomxresiduals_numbers.pdf"), width=10, height=4)
ggarrange(g9,g10, labels = c("a", "b"))
#dev.off()

```

```{r all, fig.width=10, fig.height=16, dpi=300}
pdf(paste0(sup_fig, "PCAs.pdf"), width = 10, height = 16)
ggarrange(g1,g2,g3,g4,g1,g6,g7,g8, k1, k2, ncol = 2, nrow = 5)
dev.off()
```


```{r}
sessionInfo()
```

