--
title: "Looking at covariates"
subtitle: ""
author:  |
 | Katia de Paiva Lopes 
 | Raj Lab
 | Department of Neuroscience
 | Icahn School of Medicine at Mount Sinai
 | NYC, New York
date: "`r Sys.Date()`"
output: 
 rmarkdown::html_document:
   theme: united
   highlight: tango
   code_folding: show
   toc: true
   toc_float: true
   df_print: paged
   smooth_scroll: true
   number_sections: false
   self_contained: true  
---

> 255 samples (from 314) | 
> Different brain regions from the same donor | 
> Multi disease cohort. 

Previous R script: metadata_merge.R by Katia Lopes. 

```{r clean.variables, echo=FALSE}
#This command clean all variables. BE CAREFULL!!! 
rm(list = setdiff(ls(), lsf.str()))
```

```{r load.packages, echo=FALSE, message=FALSE, results='hide'}
if(!require("variancePartition")) install.packages("variancePartition"); library("variancePartition")
if(!require("limma")) install.packages("limma"); library("limma")
if(!require("stats")) install.packages("stats"); library("stats")
if(!require("ggfortify")) install.packages("ggfortify"); library("ggfortify")
if(!require("gplots")) install.packages("gplots"); library("gplots")
if(!require("factoextra")) install.packages("factoextra"); library("factoextra")
if(!require("pamr")) BiocManager::install("pamr"); library("pamr")
if(!require("DESeq2")) BiocManager::install("DESeq2"); library("DESeq2")
if(!require("edgeR")) BiocManager::install("edgeR"); library("edgeR")
if(!require("broom")) BiocManager::install("broom"); library("broom")
if(!require("tidyverse")) BiocManager::install("tidyverse"); library("tidyverse")
library("ggeasy")
library(dplyr)
library(ggsci)
library(ggpubr)
library(ggplot2)
library(gridExtra)


knitr::opts_chunk$set( 
  warning=FALSE,
  message=FALSE,
  results = 'asis',
  error = FALSE,
  tidy = FALSE,
  fig.show = "hold")
```

```{r Helper Functions, echo=FALSE}
createDT <- function(DF, caption="", scrollY=500){
  data <- DT::datatable(DF, caption=caption,
    extensions =  'Buttons',
    options = list( dom = 'Bfrtip', 
                    buttons = c('copy', 'csv', 'excel', 'pdf', 'print'), 
                    scrollY = scrollY, scrollX=T, scrollCollapse = T, paging = F,  
                      columnDefs = list(list(className = 'dt-center', targets = "_all"))
    )
  ) 
   return(data)
}
```

```{r folders, echo=TRUE}
work_dir = "~/pd-omics/katia/Microglia/mic_255s/"
expression_dir = "~/ad-omics_hydra/microglia_omics/expression_tables/added_pilot_314s/expr_4brain_regions/"
plots4paper = "/Users/katia/OneDrive/Documentos/MountSinai/Projects/Microglia/Figures4paper/"
vp_by_tissue = "~/pd-omics/katia/Microglia/mic_255s/vp_files/"
metadata_path = "~/pd-omics/katia/Microglia/mic_255s/metadata_files/"
```

## Input: metadata
This metadata contains biological and technical covariates. 
```{r input.metadata, echo=TRUE}
load(paste0(metadata_path, "metadata_255filt_eng_29jun2020.Rdata"))
# str(metadata3rd_pass)
metadata<-metadata3rd_pass

rownames(metadata) = metadata$donor_tissue
```

## HM: correlating the covariates 
<div style="text-align: justify"> 
"Evaluating the correlation between variables in a important part in interpreting variancePartition results. When comparing two **continuous** variables, Pearson
correlation is widely used. But variancePartition includes **categorical** variables in the model as well. In order to accommodate the correlation between a continuous and a categorical variable, or two categorical variables we used canonical correlation analysis. **Canonical Correlation Analysis (CCA)** is similar to correlation between two vectors, except that CCA can accommodate matricies as well." Gabriel Hoffman. 
</div>
```{r hm_variance_partition, echo=TRUE, fig.width=10, fig.height=8, dpi=300}

# These variables have zero variance so cannot be analyzed:
#  infection_2weeks, glucocorticosteroids, batch

metadata_alt1 = metadata %>% dplyr::select(-glucocorticosteroids) # These variables have zero variance so cannot be analyzed: glucocorticosteroids
metadata_alt2 = metadata_alt1 %>% dplyr::select(-infection_2weeks) # These variables have zero variance so cannot be analyzed: infection_2weeks
metadata_alt = metadata_alt2 %>% dplyr::select(-batch) # These variables have zero variance so cannot be analyzed: batch

# names(metadata_alt) = tolower(names(metadata_alt))

# Transforms characters columns in factors columns 
indx <- sapply(metadata_alt, is.character)
metadata_alt[indx] <- lapply(metadata_alt[indx], function(x) as.factor(x))
metadata_alt$ph = as.numeric(as.character(metadata_alt$ph))
#str(metadata_alt) #shows class for all columns 
form <- as.formula(paste("", paste(colnames(metadata_alt), collapse=" + "), sep=" ~ ")) #form contains the design 
C = canCorPairs(form, metadata_alt)

# pdf(paste0(work_dir, "figures_pdf/canonical_hm.pdf"), width = 10, height = 8)
plotCorrMatrix(C, margins = c(12,18))
# dev.off()
```

## Variance Partition - 255 samples
The covariates for the formula was chosen based on canonical correlation. We need to be careful to interpret this plot because here, we are using all samples together: same individual with different brain regions. 

If the covariate have NAs in the column, we can't fit a model for Limma or Dream. The following covariates have NAs: **lane, benzodiapezines, opiates, autoimmune_diseases, smoking, infection_2weeks, alcohol_dependence_daily_use, ph, suicide_attempts and cause_of_death_categories**. 

```{r vp_alls, echo=TRUE, fig.width=8, fig.height=6, dpi=300 }
load(paste0(work_dir, "vp_files/varPart_tx_255s.RData"))

vp <- sortCols( varPart_tx )
# pdf(paste0(plots4paper, "VP_figures/vp_tec_bio_285s.pdf"), width=8, height=6)
plotVarPart(vp) +
  theme_classic() +
  easy_remove_legend() + 
  easy_rotate_x_labels(angle = 70, side = c("right"))
# dev.off()

```

```{r vp_alls_bio, echo=TRUE, fig.width=4, fig.height=4, dpi=300 }
# load(paste0(work_dir, "vp_files/varPart_tx_all.RData"))

# vp <- sortCols( varPart_tx )

vp_sel = vp[, c("donor_id", "cause_of_death_categories", "C1", "tissue", "C2", "C4", "ph","pmd_minutes", "age", "C3", "main_diagnosis", "sex", "Residuals")]

colnames(vp_sel) <- (c(paste0("Individual\n(",round(mean(vp_sel$donor_id), 4)*100,"%)"), 
          					   paste0("Cause of death\n(",round(mean(vp_sel$cause_of_death_categories), 4)*100,"%)"),
          					   paste0("Ancestry - PC1\n(",round(mean(vp_sel$C1), 4)*100,"%)"),
                       paste0("Region\n(",round(mean(vp_sel$tissue), 4)*100,"%)"), 
                       paste0("Ancestry - PC2\n(",round(mean(vp_sel$C2), 4)*100,"%)"), 
					             paste0("Ancestry - PC4\n(",round(mean(vp_sel$C4), 4)*100,"%)"),
                       paste0("pH\n(",round(mean(vp_sel$ph), 4)*100,"%)"), 
                       paste0("Post-mortem delay\n(",round(mean(vp_sel$pmd_minutes), 4)*100,"%)"), 
                       paste0("Age\n(",round(mean(vp_sel$age), 4)*100,"%)"), 
                       paste0("Ancestry - PC3\n(",round(mean(vp_sel$C3), 3)*100,"%)"),
                       paste0("Diagnosis\n(",round(mean(vp_sel$main_diagnosis), 4)*100,"%)"), 
                       paste0("Sex\n(",round(mean(vp_sel$sex), 4)*100,"%)"), 
                       paste0("Residuals\n(",round(mean(vp_sel$Residuals), 4)*100,"%)")))

# tissue.colors = pal_futurama()(8)[3:12] # Start with the color in the position 3 up to the position 12
tissue.colors = colorRampPalette(c("#1A5354FF","white"))(13)

  plotVarPart(vp_sel, ylab = "Variance explained (%)") +
  theme_classic() +
  geom_boxplot(width=0.1, outlier.colour = "black", outlier.size = 0.1, outlier.stroke = 0) +
  easy_remove_legend() + 
  scale_fill_manual(values = tissue.colors ) +
  easy_rotate_x_labels(angle = 70, side = c("right"))

```

```{r vp_alls_tec, echo=TRUE, fig.width=4, fig.height=4, dpi=300 }
# load(paste0(work_dir, "vp_files/varPart_tx_all.RData"))

# vp <- sortCols( varPart_tx )

vp_sel = vp[, c("picard_pct_mrna_bases", "picard_summed_mean", "picard_pct_ribosomal_bases", "picard_pct_pf_reads_aligned", "lane", "Residuals")]
  
colnames(vp_sel) <- (c(paste0("Pct of mRNA bases\n(",round(mean(vp_sel$picard_pct_mrna_bases), 4)*100,"%)"), 
                       paste0("Mean Insert size\n(",round(mean(vp_sel$picard_summed_mean), 4)*100,"%)"), 
                       paste0("Pct of ribosome\n(",round(mean(vp_sel$picard_pct_ribosomal_bases), 4)*100,"%)"), 
                       paste0("Pct of aligned reads\n(",round(mean(vp_sel$picard_pct_pf_reads_aligned), 4)*100,"%)"), 
                       paste0("Lane\n(",round(mean(vp_sel$lane), 4)*100,"%)"), 
                       paste0("Residuals\n(",round(mean(vp_sel$Residuals), 4)*100,"%)")))

# tissue.colors = c("#008EA0FF", "#8A4198FF", "#5A9599FF", "#FF6348FF", "#84D7E1FF",  NA)  
tissue.colors = colorRampPalette(c("#ADE2D0FF","white"))(6)

plotVarPart(vp_sel, ylab = "Variance explained (%)") +
  theme_classic() +
  geom_boxplot(width=0.1, outlier.colour = "black", outlier.size = 0.1, outlier.stroke = 0) +
  easy_remove_legend() + 
  scale_fill_manual(values = tissue.colors ) +
  easy_rotate_x_labels(angle = 70, side = c("right"))

```

## Loading VP data by region 
### MFG
```{r load_by_tissue, fig.width=6, fig.height=4}
load(paste0(vp_by_tissue, "varPart_tx_MFG.RData"))

vp <- sortCols( varPart_tx )

colnames(vp) <- (c(paste0("Pct of mRNA bases\n(",round(mean(vp$picard_pct_mrna_bases), 4)*100,"%)"),
				   paste0("Ancestry - PC1\n(",round(mean(vp$C1), 4)*100,"%)"),
				   paste0("Mean Insert size\n (",round(mean(vp$picard_summed_mean), 4)*100,"%)"),
				   paste0("Ancestry - PC2\n(",round(mean(vp$C2), 4)*100,"%)"), 
				   paste0("Pct of aligned reads\n(",round(mean(vp$picard_pct_pf_reads_aligned), 4)*100,"%)"),
				   paste0("Pct of ribosome\n(",round(mean(vp$picard_pct_ribosomal_bases), 4)*100,"%)"),
				   paste0("Ancestry - PC4\n(",round(mean(vp$C4), 3)*100,"%)"),
				   paste0("pH\n(",round(mean(vp$ph), 4)*100,"%)"), 
				   paste0("Lane\n(",round(mean(vp$lane), 4)*100,"%)"),
				   paste0("Ancestry - PC3\n(",round(mean(vp$C3), 3)*100,"%)"),
				   paste0("Age\n(",round(mean(vp$age), 4)*100,"%)"),
				   paste0("Post-mortem delay\n(",round(mean(vp$pmd_minutes), 4)*100,"%)"), 
				   paste0("Cause of death\n(",round(mean(vp$cause_of_death_categories), 4)*100,"%)"),
				   paste0("Sex\n(",round(mean(vp$sex), 4)*100,"%)"), 
				   paste0("Diagnosis\n(",round(mean(vp$main_diagnosis), 4)*100,"%)"), 
				   paste0("Residuals\n(",round(mean(vp$Residuals), 4)*100,"%)")))

tissue.colors = colorRampPalette(c("#FF6F00FF","white"))(16)

p1 <- plotVarPart(vp, main = "MFG") +
  theme_classic() +
  geom_boxplot(width=0.1, outlier.colour = "black", outlier.size = 0.1, outlier.stroke = 0) +
  easy_remove_legend() +
  scale_fill_manual(values = tissue.colors ) +
  easy_rotate_x_labels(angle = 70, side = c("right"))

```

### STG
```{r load_by_tissue, fig.width=6, fig.height=4}
load(paste0(vp_by_tissue, "varPart_tx_STG.RData"))

vp <- sortCols( varPart_tx )

colnames(vp) <- (c(paste0("Pct of mRNA bases\n(",round(mean(vp$picard_pct_mrna_bases), 4)*100,"%)"),
				   paste0("Ancestry - PC1\n(",round(mean(vp$C1), 4)*100,"%)"),
				   paste0("Ancestry - PC2\n(",round(mean(vp$C2), 4)*100,"%)"),
				   paste0("Mean Insert size\n (",round(mean(vp$picard_summed_mean), 4)*100,"%)"),
				   paste0("Pct of aligned reads\n(",round(mean(vp$picard_pct_pf_reads_aligned), 4)*100,"%)"),
				   paste0("Pct of ribosome\n(",round(mean(vp$picard_pct_ribosomal_bases), 4)*100,"%)"),
				   paste0("Ancestry - PC4\n(",round(mean(vp$C4), 3)*100,"%)"),
				   paste0("Ancestry - PC3\n(",round(mean(vp$C3), 3)*100,"%)"),
				   paste0("pH\n(",round(mean(vp$ph), 4)*100,"%)"), 
				   paste0("Post-mortem delay\n(",round(mean(vp$pmd_minutes), 4)*100,"%)"), 
				   paste0("Age\n(",round(mean(vp$age), 4)*100,"%)"),
				   paste0("Lane\n(",round(mean(vp$lane), 4)*100,"%)"),
				   paste0("Cause of death\n(",round(mean(vp$cause_of_death_categories), 4)*100,"%)"),
				   paste0("Sex\n(",round(mean(vp$sex), 4)*100,"%)"), 
				   paste0("Diagnosis\n(",round(mean(vp$main_diagnosis), 4)*100,"%)"), 
				   paste0("Residuals\n(",round(mean(vp$Residuals), 4)*100,"%)")))

tissue.colors = colorRampPalette(c("#C71000FF","white"))(16)

p2 <- plotVarPart(vp, main = "STG") +
  theme_classic() +
  geom_boxplot(width=0.1, outlier.colour = "black", outlier.size = 0.1, outlier.stroke = 0) +
  easy_remove_legend() +
  scale_fill_manual(values = tissue.colors ) +
  easy_rotate_x_labels(angle = 70, side = c("right"))

# colorRampPalette(c("#C71000FF","white"))(6)
# ggarrange(p1,p2)
```

```{r load_by_tissue, fig.width=6, fig.height=4}
load(paste0(vp_by_tissue, "varPart_tx_THA.RData"))

vp <- sortCols( varPart_tx )

colnames(vp) <- (c(paste0("Pct of mRNA bases\n(",round(mean(vp$picard_pct_mrna_bases), 4)*100,"%)"),
				   paste0("Ancestry - PC1\n(",round(mean(vp$C1), 4)*100,"%)"),
				   paste0("Pct of aligned reads\n(",round(mean(vp$picard_pct_pf_reads_aligned), 4)*100,"%)"),
				   paste0("Mean Insert size\n (",round(mean(vp$picard_summed_mean), 4)*100,"%)"),
				   paste0("Ancestry - PC2\n(",round(mean(vp$C2), 4)*100,"%)"),
				   paste0("Pct of ribosome\n(",round(mean(vp$picard_pct_ribosomal_bases), 4)*100,"%)"),
				   paste0("Cause of death\n(",round(mean(vp$cause_of_death_categories), 4)*100,"%)"),
				   paste0("Ancestry - PC3\n(",round(mean(vp$C3), 3)*100,"%)"),
				   paste0("Ancestry - PC4\n(",round(mean(vp$C4), 3)*100,"%)"),
				   paste0("Post-mortem delay\n(",round(mean(vp$pmd_minutes), 4)*100,"%)"),
				   paste0("Age\n(",round(mean(vp$age), 4)*100,"%)"),
				   paste0("pH\n(",round(mean(vp$ph), 4)*100,"%)"), 
				   paste0("Sex\n(",round(mean(vp$sex), 4)*100,"%)"),
				   paste0("Lane\n(",round(mean(vp$lane), 4)*100,"%)"),
				   paste0("Diagnosis\n(",round(mean(vp$main_diagnosis), 4)*100,"%)"), 
				   paste0("Residuals\n(",round(mean(vp$Residuals), 4)*100,"%)")))

tissue.colors = colorRampPalette(c("#8A4198FF","white"))(16)

p3 <- plotVarPart(vp, main = "THA") +
  theme_classic() +
  geom_boxplot(width=0.1, outlier.colour = "black", outlier.size = 0.1, outlier.stroke = 0) +
  easy_remove_legend() +
  scale_fill_manual(values = tissue.colors ) +
  easy_rotate_x_labels(angle = 70, side = c("right"))

colorRampPalette(c("#8A4198FF","white"))(6)

```

## SVZ 
```{r load_by_tissue, fig.width=6, fig.height=4}
load(paste0(vp_by_tissue, "varPart_tx_SVZ.RData"))

vp <- sortCols( varPart_tx )

colnames(vp) <- (c(paste0("Pct of mRNA bases\n(",round(mean(vp$picard_pct_mrna_bases), 4)*100,"%)"),
				   paste0("Ancestry - PC1\n(",round(mean(vp$C1), 4)*100,"%)"),
				   paste0("Pct of aligned reads\n(",round(mean(vp$picard_pct_pf_reads_aligned), 4)*100,"%)"),
				   paste0("Mean Insert size\n (",round(mean(vp$picard_summed_mean), 4)*100,"%)"),
				   paste0("Ancestry - PC2\n(",round(mean(vp$C2), 4)*100,"%)"),
				   paste0("Pct of ribosome\n(",round(mean(vp$picard_pct_ribosomal_bases), 4)*100,"%)"),
				   paste0("Ancestry - PC4\n(",round(mean(vp$C4), 3)*100,"%)"),
				   paste0("Ancestry - PC3\n(",round(mean(vp$C3), 3)*100,"%)"),
				   paste0("Age\n(",round(mean(vp$age), 4)*100,"%)"),
				   paste0("Post-mortem delay\n(",round(mean(vp$pmd_minutes), 4)*100,"%)"),
				   paste0("pH\n(",round(mean(vp$ph), 4)*100,"%)"), 
				   paste0("Cause of death\n(",round(mean(vp$cause_of_death_categories), 4)*100,"%)"),
				   paste0("Lane\n(",round(mean(vp$lane), 4)*100,"%)"),
				   paste0("Sex\n(",round(mean(vp$sex), 4)*100,"%)"),
				   paste0("Diagnosis\n(",round(mean(vp$main_diagnosis), 4)*100,"%)"), 
				   paste0("Residuals\n(",round(mean(vp$Residuals), 4)*100,"%)")))

tissue.colors = colorRampPalette(c("#008EA0FF","white"))(16) 

p4 <- plotVarPart(vp, main = "SVZ") +
  theme_classic() +
  geom_boxplot(width=0.1, outlier.colour = "black", outlier.size = 0.1, outlier.stroke = 0) +
  easy_remove_legend() +
  scale_fill_manual(values = tissue.colors ) +
  easy_rotate_x_labels(angle = 70, side = c("right"))

colorRampPalette(c("#008EA0FF","white"))(6)

```

```{r load_by_tissue, fig.width=6, fig.height=4, eval=FALSE}
load(paste0(vp_by_tissue, "varPart_tx_CC.RData"))

vp <- sortCols( varPart_tx )

colnames(vp) <- (c(paste0("Pct of mRNA bases\n(",round(mean(vp$picard_pct_mrna_bases), 4)*100,"%)"),
                   paste0("Mean Insert size\n (",round(mean(vp$picard_summed_mean), 4)*100,"%)"),
                   paste0("Pct of aligned reads\n(",round(mean(vp$picard_pct_pf_reads_aligned), 4)*100,"%)"),
                   paste0("pH\n(",round(mean(vp$ph), 4)*100,"%)"), 
                   paste0("Pct of ribosome\n(",round(mean(vp$picard_pct_ribosomal_bases), 4)*100,"%)"),
                   paste0("Age\n(",round(mean(vp$age), 4)*100,"%)"), 
                   paste0("Cause of death\n(",round(mean(vp$cause_of_death_categories), 4)*100,"%)"), 
                   paste0("Diagnosis\n(",round(mean(vp$main_diagnosis), 4)*100,"%)"), 
                   paste0("Sex\n(",round(mean(vp$sex), 4)*100,"%)"), 
                   paste0("Residuals\n(",round(mean(vp$Residuals), 4)*100,"%)")))

p5 <- plotVarPart(vp, main = "CC") +
  theme_classic() +
  geom_boxplot(width=0.1, outlier.colour = "black", outlier.size = 0.1, outlier.stroke = 0) +
  easy_remove_legend() +
  scale_fill_manual(values = c("#F493C0", "#F493C0", "#F493C0", "#F493C0","#FF95A8FF", "#EE4C97FF", "#EE4C97FF", "#EE4C97FF", "#EE4C97FF", "white")) +
  #scale_fill_manual(values = tissue.colors ) +
  easy_rotate_x_labels(angle = 70, side = c("right"))

colorRampPalette(c("#EE4C97FF","white"))(6)

```

```{r fig.width=12, fig.height=8 }
ggarrange(p1,p2,p3,p4, ncol=2, nrow = 2, labels = c("a", "b", "c", "d"))

```

