---
title: "Aging analysis for sQTL"
subtitle: " "
author:  |
 | Katia de Paiva Lopes 
 | Jack Humphrey
 | Raj Lab
 | Department of Neuroscience
 | Icahn School of Medicine at Mount Sinai
 | NYC, New York
date: "`r Sys.Date()`"
output: 
 rmarkdown::html_document:
   theme: united
   highlight: tango
   code_folding: hide
   toc: true
   toc_float: true
   df_print: paged
   smooth_scroll: true
   number_sections: false
   self_contained: true  
---

> 255 samples | 
> Transcripts from RSEM | 
> Dream | coef = age.   

```{r clean.variables, echo=FALSE}
#This command clean all variables. BE CAREFULL!!! 
rm(list = setdiff(ls(), lsf.str()))

library(edgeR)
library(ggplot2)
library('doParallel')
library("BiocParallel")
library('tidyverse')
library("variancePartition")

knitr::opts_chunk$set( 
  warning=FALSE,
  message=FALSE,
  results = 'asis',
  error = FALSE,
  tidy = FALSE,
  fig.show = "hold")
```

```{r Helper Functions, echo=FALSE}
createDT <- function(DF, caption="", scrollY=500){
  data <- DT::datatable(DF, caption=caption,
    extensions =  'Buttons',
    options = list( dom = 'Bfrtip', 
                    buttons = c('copy', 'csv', 'excel', 'pdf', 'print'), 
                    scrollY = scrollY, scrollX=T, scrollCollapse = T, paging = F,  
                      columnDefs = list(list(className = 'dt-center', targets = "_all"))
    )
  ) 
   return(data)
}
```

```{r path_data}
expression_path = "~/ad-omics_hydra/microglia_omics/expression_tables/added_pilot_314s/"
metadata_path = "~/pd-omics/katia/Microglia/mic_255s/metadata_files/"
```

```{r input_data}
load(paste0(metadata_path, "metadata_255filt_eng_29jun2020.Rdata"))
# str(metadata3rd_pass)

load(paste0(expression_path, "tx_matrix.RData"))
# dim(tx_counts)
# dim(tx_tpm)

tx_counts_filt = tx_counts
colnames(tx_counts_filt) = gsub("(.*?)\\.(.*)", "\\1", colnames(tx_counts_filt))

tx_counts_filt = tx_counts_filt[, as.character(metadata3rd_pass$donor_tissue)]

#Filter genes:
x = data.frame(tx_counts_filt, check.names = F)
cpm = cpm(x)
keep.exp = rowSums(cpm > 1) >= 0.3*ncol(x) 
tx_counts_filt = tx_counts_filt[keep.exp,] # final filtered transcript table

#save(tx_counts_filt, file = paste0(expression_path, "/expr_255s_tx/tx_matrix_255s_filt.RData"))
```

## Age distribution
### 255 samples 
```{r age_plots, echo=TRUE, fig.width=4, fig.height=3}
metadata <- metadata3rd_pass
ageByDonor = unique(metadata[,c("donor_id", "age", "sex")])

ggplot(ageByDonor, aes(x=age, fill=age)) +
  geom_histogram(bins = 25, colour='black', position = "stack", fill="#5c88da99") +
  labs(x="Age", y="Donors") +
  scale_y_continuous(breaks = (1:20)) +
  scale_x_continuous(breaks=seq(20,120,10)) + 
  theme_classic()

```

## Dream analysis {.tabset .tabset-fade .tabset-pills} 
To use **cause_of_death** in the model, I got the NAs and change it for "Other" category. 

To use C1-C4 I calculated the median for the missing samples. 
```{r dream, echo=TRUE, eval=F}
params = BiocParallel::MulticoreParam(workers=20, progressbar=T)
register(params)
registerDoParallel(20)

metadata$cause_of_death_categories[metadata$cause_of_death_categories %in% NA] <- "Other"
# table(metadata$cause_of_death_categories)

metadata$C1 = metadata$C1 %>% replace_na(median(metadata$C1, na.rm = T))
metadata$C2 = metadata$C2 %>% replace_na(median(metadata$C2, na.rm = T))
metadata$C3 = metadata$C3 %>% replace_na(median(metadata$C3, na.rm = T))
metadata$C4 = metadata$C4 %>% replace_na(median(metadata$C4, na.rm = T))

# Matching the names from the DE analysis 
genes_counts4deg = tx_counts_filt
metadata4deg = metadata
# all(colnames(tx_counts_filt) == as.character(metadata4deg$donor_tissue)) # Check the order of columns - TRUE 

# The dream model operates directly on the results of voom. 
# The only change compared to the standard limma workflow is to replace lmFit with dream. 

# Check variance partition version 
# packageVersion("variancePartition")  # Must be 1.17.7

# The variable to be tested should be a fixed effect
form <- ~ sex + (1|donor_id) + age + tissue + (1|cause_of_death_categories) + C1 + C2 + C3 + C4 + picard_pct_mrna_bases + picard_summed_median + picard_pct_ribosomal_bases

# estimate weights using linear mixed model of dream
vobjDream = suppressWarnings( voomWithDreamWeights( genes_counts4deg, form, metadata4deg ) ) # supressing messages because of Biocparallel was generating a lot of messages  
 
# Fit the dream model on each gene
# By default, uses the Satterthwaite approximation for the hypothesis test
fitmm = suppressWarnings (dream( vobjDream, form, metadata4deg )) 

# Examine design matrix
# createDT(fitmm$design, 3)

res_age <- data.frame(topTable(fitmm, coef='age', 
                                 number=nrow(genes_counts4deg), sort.by = "p"), check.names = F)

save(vobjDream, form, fitmm, res_age, genes_counts4deg, metadata4deg, file = "dream_DE_res.RData")
```

## Transcripts related with age

```{r design0.lists, echo=TRUE}
load("dream_DE_res.RData")
res = res_age
res$ensembl_transcript = rownames(res)

## Get conversion table for Gencode 30
gencode_30 = read.table("~/pd-omics/katia/ens.geneid.gencode.v30", header = T)
colnames(gencode_30) = c("ensembl","symbol")
gencode_30_map = read.table("~/pd-omics/katia/gencode.v30.primary_assembly.annotation.reflat", header = F)[,1:2]
colnames(gencode_30_map) = c("ensembl","ensembl_transcript")
gencode_30_map = gencode_30_map %>% left_join(gencode_30)

res_name = merge(res, gencode_30_map, by="ensembl_transcript")
rownames(res_name) = res_name$ensembl_transcript
res_name = res_name[order(res_name$adj.P.Val), ]
res_name = res_name[, c("symbol", "logFC", "AveExpr", "t", "P.Value", "adj.P.Val")]
   
#deg_lists = res_name[which(res_name$adj.P.Val<0.001),]

write.table(res_name, file = paste0("List_age_dream_255s.txt"), quote = F, sep = "\t")
# write.table(deg_lists, file = paste0(work_plots, "List_age_0.001_dream.txt"), quote = F, sep = "\t")

createDT(res_name)
```

### adj.P.Val<0.001
```{r design0.deg1, echo=TRUE, results='asis'}
print(paste0("Num DE transcripts = ", length(which(res_name$adj.P.Val<0.001))))
print(paste0("Num DE unique genes = ", nrow(res_name %>% filter(adj.P.Val<0.001) %>% select(symbol) %>% distinct())))
```

### adj.P.Val<0.01
```{r design0.deg2, echo=TRUE, results='asis'}
print(paste0("Num DE transcripts = ", length(which(res_name$adj.P.Val<0.01))))
print(paste0("Num DE unique genes = ", nrow(res_name %>% filter(adj.P.Val<0.01) %>% select(symbol) %>% distinct())))
```

### adj.P.Val<0.05
```{r design0.deg3, echo=TRUE, results='asis'}
print(paste0("Num DE transcripts = ", length(which(res_name$adj.P.Val<0.05))))
print(paste0("Num DE unique genes = ", nrow(res_name %>% filter(adj.P.Val<0.05) %>% select(symbol) %>% distinct())))
```


## PCA voom 

The voom here comes from the Dream object. 

```{r pca_voom, echo=TRUE, fig.width=6, fig.height=4}
genes_voom = as.data.frame(vobjDream$E) 
res.pca = prcomp(t(genes_voom)) 
library(ggfortify); library(ggplot2)
autoplot(res.pca, data = metadata4deg, colour = 'age') +
  scale_colour_viridis_c() +
  theme_classic()

```

I'm applying the removeBatchEffect function to get the residuals. 
```{r remove_batch, echo=TRUE}
allResiduals <- removeBatchEffect(x = genes_voom, 
                                  batch = metadata4deg$tissue, 
                                  batch2 = paste(metadata4deg$sex, metadata4deg$cause_of_death_categories),
                                 # design = model.matrix(~ age, data = metadata4deg), #force to not regress age. It didn't change the PCA. 
                                  covariates = as.matrix(metadata4deg[, c("picard_pct_mrna_bases", "picard_summed_median", "picard_pct_ribosomal_bases", "C1","C2","C3","C4" )]))

```

## PCA after removeBatchEffect
```{r pca_noBatch, echo=TRUE, fig.width=6, fig.height=4}
res.pca = prcomp(t(allResiduals)) 

autoplot(res.pca, data = metadata4deg, colour = 'age') +
  scale_colour_viridis_c() +
  theme_classic()
```

```{r}
sessionInfo()
```


