---
title: "Post mashR pipeline"
subtitle: "sQTL"
author:  |
 | Katia de Paiva Lopes 
 | Raj Lab
 | Department of Neuroscience
 | Icahn School of Medicine at Mount Sinai
 | NYC, New York
date: "`r Sys.Date()`"
output: 
 rmarkdown::html_document:
   theme: united
   highlight: tango
   code_folding: hide
   toc: true
   toc_float: true
   df_print: paged
   smooth_scroll: true
   number_sections: false
   self_contained: true  
---

> Previous pipeline: https://github.com/RajLabMSSM/mashR 

> GTEx paper: https://www.nature.com/articles/s41588-018-0268-8 

```{r clean.variables, echo=FALSE}
#This command clean all variables. BE CAREFULL!!! 
rm(list = setdiff(ls(), lsf.str()))
```

```{r load.packages, echo=FALSE, message=FALSE, results='hide'}
library("lattice")
library("dplyr")
library("rmeta")
library("ggsci")

fastqtl_to_mash_output = "~/pd-omics/katia/Microglia/mic_255s/mashr_sqtl_3rd/fastqtl_to_mash_output/"
mashr_flashr_workflow_output = "~/pd-omics/katia/Microglia/mic_255s/mashr_sqtl_3rd/mashr_flashr_workflow_output/"
code_folder = "~/ad-omics_hydra/microglia_omics/mash_sQTL_no2points_id_OLD/code/"
work_dir = "~/pd-omics/katia/Microglia/mic_255s/mashr_sqtl_3rd/"
  
knitr::opts_chunk$set( 
  warning=FALSE,
  message=FALSE,
  results = 'asis',
  error = FALSE,
  tidy = FALSE,
  fig.show = "hold")

```

```{r Helper Functions, echo=FALSE}
createDT <- function(DF, caption="", scrollY=500){
  data <- DT::datatable(DF, caption=caption,
    extensions =  'Buttons',
    options = list( dom = 'Bfrtip', 
                    buttons = c('copy', 'csv', 'excel', 'pdf', 'print'), 
                    scrollY = scrollY, scrollX=T, scrollCollapse = T, paging = F,  
                      columnDefs = list(list(className = 'dt-center', targets = "_all"))
    )
  ) 
   return(data)
}
```

```{r input_data}
mash_rds = readRDS(paste0(fastqtl_to_mash_output, "QTLSumStats.mash.rds"))
#names(mash_rds)

maxb <- mash_rds$strong.b
maxz <- mash_rds$strong.z

mash_posterior = readRDS(paste0(mashr_flashr_workflow_output, "QTLSumStats.mash.EZ.FL_PC3.V_simple.posterior.rds"))
# names(mash_posterior)
# head(mash_posterior$lfsr)
# dim(mash_posterior$lfsr)

pm.mash <- mash_posterior$PosteriorMean
lfsr.all <- mash_posterior$lfsr
standard.error <- maxb/maxz
pm.mash.beta <- pm.mash*standard.error
```

## lfsr table at 5%
lfsr = local false sign rate. Method proposed by Stephens, it is analogous to FDR. 

Number of significative sclusters:
```{r lfsr.sign}
pm.mash.beta <- pm.mash.beta[rowSums(lfsr.all<0.05)>0,]
lfsr.mash <- lfsr.all[rowSums(lfsr.all<0.05)>0,] # lfsr.mash have the significant results. 
# Means they have lfsr less than 0.05 in at least one condition.
dim(lfsr.mash)[1] 
lfsr.mash_symbol = as.data.frame(lfsr.mash)
rownames(lfsr.mash_symbol) = gsub("_ENS", ":ENS", rownames(lfsr.mash_symbol))
lfsr.mash_symbol$ensembl = gsub("(.*?):(.*)", "\\2", rownames(lfsr.mash_symbol))
lfsr.mash_symbol$ensembl = gsub("(.*?)_(.*)", "\\1", lfsr.mash_symbol$ensembl)

## Get conversion table for Gencode 30
gencode_30 = read.table("~/pd-omics/katia/ens.geneid.gencode.v30", header = T)
gencode_30$ensembl = gsub("(.*?)\\.(.*)", "\\1", gencode_30$gene_id)
lfsr.mash_symbol$scluster = rownames(lfsr.mash_symbol)
lfsr.mash_symbol = merge(lfsr.mash_symbol, gencode_30, by = "ensembl")
colnames(lfsr.mash_symbol) = gsub("(.*?)_(.*)","\\1",colnames(lfsr.mash_symbol))

lfsr.mash_symbol$gene_id <- NULL 
lfsr.mash_symbol = lfsr.mash_symbol[,c("ensembl", "GeneSymbol", "scluster", "MFG", "STG", "SVZ", "THA")]
lfsr.mash_symbol$scluster = gsub("_s", "_rs", lfsr.mash_symbol$scluster)

write.table(lfsr.all, file = paste0(work_dir, "mash_lfsr_sQTL_all.txt"), sep = "\t", quote = F)
write.table(lfsr.mash_symbol, file = paste0(work_dir, "mash_lfsr_sQTL_5per.txt"), sep = "\t", quote = F)

createDT(lfsr.mash_symbol)
```

Number of unique gene:
```{r uniq_gene}
length(unique(lfsr.mash_symbol$GeneSymbol))
```

## Pairwise plot

<div style="text-align: justify"> 
Pairwise sharing by magnitude of sQtLs among tissues. For each pair of tissues, we considered the top sQTLs that were significant (lfsr < 0.05) in at least one of the two tissues, and plotted the proportion of these that are shared in magnitude—that is, have effect estimates that are the same sign and within a factor of 2 in size of one another.
</div>
```{r heatmap, fig.height=4, fig.width=4, dpi=300}
thresh <- 0.05

shared.fold.size <- matrix(NA,nrow = ncol(lfsr.mash),ncol=ncol(lfsr.mash))
colnames(shared.fold.size) <- rownames(shared.fold.size) <- colnames(maxz)
for (i in 1:ncol(lfsr.mash))
  for (j in 1:ncol(lfsr.mash)) {
    sig.row=which(lfsr.mash[,i]<thresh)
    sig.col=which(lfsr.mash[,j]<thresh)
    a=(union(sig.row,sig.col))
    quotient=(pm.mash.beta[a,i]/pm.mash.beta[a,j])
    shared.fold.size[i,j] = mean(quotient > 0.5 & quotient < 2)
  }

# Plot heatmap of sharing by magnitude
# Generate the heatmap using the “levelplot” function from the lattice package.

clrs <- colorRampPalette(rev(c("#D73027","#FC8D59","#FEE090","#FFFFBF",
                               "#E0F3F8","#91BFDB","#4575B4")))(64)
lat <- shared.fold.size
lat[lower.tri(lat)] <- NA
n <- nrow(lat)

rownames(lat) <- gsub("(.*?)_(.*)", "\\1", rownames(lat))
colnames(lat) <- gsub("(.*?)_(.*)", "\\1", colnames(lat))

myPanel <- function(x, y, z, ...) {
  panel.levelplot(x,y,z,...)
  panel.text(x, y, ifelse(is.na(round(z,2)),"",round(z,2)))
}

#pdf(file = "/Users/katia/OneDrive/Documentos/MountSinai/Projects/Microglia/Figures4paper/pairwise_numb_005_sQTL.pdf", width = 5, height = 5)
print(levelplot( lat[n:1,],col.regions = clrs,xlab = "",ylab = "",colorkey = TRUE, panel=myPanel,
                 main = "Pairwise sharing by magnitude"))
#dev.off()

```

## Sharing by sign 
Compute overall sharing by sign and magnitude 
```{r overall_shar}
source(paste0(code_folder, "normfuncs.R"))

# Sharing by sign 
signall <- mean(het.norm(pm.mash.beta)>0)
signall #sQTL = 97 % # eQTL = 89 % 
```

## Sharing by magnitude 
```{r overall_shar2}
# Sharing by magnitude 
magall <- mean(het.norm(pm.mash.beta)>0.5) 
magall #sQTL = 78%  # eQTL = 53%
```

```{r}
sessionInfo()
```

